---
title: "PlatypusV3 EAE"
author: "Victor Kreiner, LSSI at D-BSSE"
date: "22 9 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PlatypusV3 EAE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette") #not necessary

source_code_dir <- "C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/PlatypusDB_dev_VK/R" 
file_path_vec <- list.files(source_code_dir, full.names = T)
for(f_path in file_path_vec){
  print(f_path)
  tryCatch({source(f_path)}, error = function(e){e})
}
gc()

#to avoid of the dplyr summarise warnings
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)
library(stats, warn.conflicts = F)

knitr::opts_chunk$set(fig.width=6, fig.height=6) 
```

## 1. Introduction

Adaptive immune phenotypes and repertoires in EAE - a full Platypus V3 walkthrough

This vignette demonstrates the functionality of the Platypus v3 analysis pipeline.  

For any questions or suggestions, please reach out to myself (Victor Kreiner, vkreiner@campus.lmu.de) or Alexander Yermanos, ayermanos@gmail.com)

## 2. Dataset overview 

Recent developments in single-cell immune repertoire sequencing have enabled the integration of lymphocyte’s transcriptional phenotypes and their adaptive immune receptors. This provides a systems-level view of an ongoing immune response. Here we performed paired single-cell RNA and immune repertoire receptor sequencing of T and B cells in the mouse model of experimental autoimmune encephalomyelitis (EAE). The aim was to explore the clonal selection experienced by lymphocytes and their immune receptors in this condition. B and T cells were isolated from the brains of B6 mice 14 days following induction of EAE with either rMOG protein (1-125) or peptide (35-55).

## 3. Introduction to the analysis pipeline

The majority of analyses in this vignette use Platypus v3, which revolves around the VDJ_GEX_matrix (VGM) data structure. This is used as input to the majority of downstream analysis functions.

Data will be read in locally, but an alternative read in command from PlatypusDB for all data is provided as well. 
Many lines are commented out for the purpose of vignette compilation (e.g. #ggsave() calls).
The VDJ_GEX_matrix function will be run multiple times for the purpose of integration and filtering. We highly suggest to save the output objects as .rds via saveRDS() in order to avoid running this function each time when starting a new R session.

## 4. Setup 

Loading in the needed packages. The pals and scales packages will be used for color paletts
We further load the ggrepel package explicitly to shorten ggplot commands. 

```{r, fig.show='hold', message = FALSE, warning=FALSE}
library(Platypus)
library(Seurat)
library(tidyverse)
library(utils)
library(scales)
library(pals)
library(ggrepel)
library(igraph)
```


```{r, fig.show='hold', message = FALSE, include=FALSE}
source_code_dir <- "C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/PlatypusDB_dev_VK/R" 
file_path_vec <- list.files(source_code_dir, full.names = T)
for(f_path in file_path_vec){
  print(f_path)
  tryCatch({source(f_path)}, error = function(e){e})
}
gc()

```

## Import and VGM runs

Import locally

```{r, fig.show='hold'}
VDJ.out.directory.list_T <- list()
VDJ.out.directory.list_T[[1]] <- c("~/EAE/tcell/S1") #rMOG_1 
VDJ.out.directory.list_T[[2]] <- c("~/EAE/tcell/S2") #rMOG_2
VDJ.out.directory.list_T[[3]] <- c("~/EAE/tcell/S3") #MOG35-55

VDJ.out.directory.list_B <- list()
VDJ.out.directory.list_B[[1]] <- c("~/EAE/bcell/S1") #rMOG_1 
VDJ.out.directory.list_B[[2]] <- c("~/EAE/bcell/S2") #rMOG_2
VDJ.out.directory.list_B[[3]] <- c("~/EAE/bcell/S3") #MOG35-55

GEX.out.directory.list <- list()
GEX.out.directory.list[[1]] <- c("~/EAE/bcell/S1") #rMOG_1 
GEX.out.directory.list[[2]] <- c("~/EAE/bcell/S2") #rMOG_2
GEX.out.directory.list[[3]] <- c("~/EAE/bcell/S3") #MOG35-55
```

```{r, include = F}
VDJ.out.directory.list_T <- list()
#EAE
VDJ.out.directory.list_T[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/tcell/S1") #Protein acute brain + spinal pooled
VDJ.out.directory.list_T[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/tcell/S2") #Protein acute brain + spinal pooled
VDJ.out.directory.list_T[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/tcell/S3") #Peptide acute brain + spinal

VDJ.out.directory.list_B <- list()
#EAE
VDJ.out.directory.list_B[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S1") #Protein acute brain + spinal pooled
VDJ.out.directory.list_B[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S2") #Protein acute brain + spinal pooled
VDJ.out.directory.list_B[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S3") #Peptide acute brain + spinal

GEX.out.directory.list <- list()
#EAE
GEX.out.directory.list[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S1") #Protein acute brain + spinal pooled
GEX.out.directory.list[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S2") #Protein acute brain + spinal pooled
GEX.out.directory.list[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S3") #Peptide acute brain + spinal

vgm_b <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_b.rds")
vgm_t <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_t.rds")
vgm_allgex <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_allgex.rds")

```

Import from PlatypusDB

```{r, fig.show='hold'}
EAE_Tcells <- PlatypusDB_fetch(
              PlatypusDB.links = c("kreiner2021a/ALL/ALL"),
              save.to.disk = F, 
              load.to.enviroment = F, 
              load.to.list = T, 
              combine.objects = T)

EAE_Bcells <- PlatypusDB_fetch(
              PlatypusDB.links = c("kreiner2021b/ALL/ALL"),
              save.to.disk = F, 
              load.to.enviroment = F, 
              load.to.list = T, 
              combine.objects = T)
```

# Run VGM

From local data

```{r, fig.show='hold', eval = FALSE}
#No GEX filter => ALL GEX + T cell repertoire
vgm_allgex <- VDJ_GEX_matrix(VDJ.out.directory.list = VDJ.out.directory.list_T,                         
                           GEX.out.directory.list = GEX.out.directory.list,
                           exclude.on.cell.state.markers = c("no filtering"), 
                           group.id = c("rMOG","rMOG","MOG35-55")) 
saveRDS(vgm_allgex,"EAE_vgm_allgex.rds")

#T cell GEX + T cell VDJ
vgm_t <- VDJ_GEX_matrix(VDJ.out.directory.list = VDJ.out.directory.list_T,                         
                           GEX.out.directory.list = GEX.out.directory.list,
                           exclude.on.cell.state.markers = c("CD19+", "CD3E-", "EBF1+"), 
                           group.id = c("rMOG","rMOG","MOG35-55")) 
saveRDS(vgm_t,"EAE_vgm_t.rds")

#B cell GEX and B cell VDJ
vgm_b <- VDJ_GEX_matrix(VDJ.out.directory.list = VDJ.out.directory.list_B,                         
                           GEX.out.directory.list = GEX.out.directory.list,
                           exclude.on.cell.state.markers = c("CD3G+", "CD3E+", "CD4+", "CD8A+","CD8B1+","C1QA+","C1QB+","RORA+","NKG7+"), #Strict filtering to achieve best possible clustering of B cells
                           group.id = c("rMOG","rMOG","MOG35-55")) 
saveRDS(vgm_b,"EAE_vgm_b.rds")

#Data architecture and sample counts
names(vgm_allgex[[1]]) #VDJ
names(vgm_allgex[[2]]@meta.data) #GEX
table(vgm_allgex[[1]]$sample_id) #Nr of cells per sample
```

From downloaded data

```{r, fig.show='hold', eval = FALSE}
#No GEX filter => ALL GEX + T cell repertoire
vgm_allgex <- VDJ_GEX_matrix(Data.in = EAE_Tcells,                         
                             exclude.on.cell.state.markers = c("no filtering"), 
                             group.id = c("rMOG","rMOG","MOG35-55")) 
saveRDS(vgm_allgex,"EAE_vgm_allgex.rds")

#T cell GEX + T cell VDJ
vgm_t <- VDJ_GEX_matrix(Data.in = EAE_Tcells
                        exclude.on.cell.state.markers = c("CD19+", "CD3E-", "EBF1+"), 
                        group.id = c("rMOG","rMOG","MOG35-55")) 
saveRDS(vgm_t,"EAE_vgm_t.rds")

#B cell GEX and B cell VDJ
vgm_b <- VDJ_GEX_matrix(Data.in = EAE_Bcells
                        exclude.on.cell.state.markers = c("CD3G+", "CD3E+", "CD4+", "CD8A+","CD8B1+","C1QA+","C1QB+","RORA+","NKG7+"), #Strict filtering to achieve best possible clustering of B cells
                       group.id = c("rMOG","rMOG","MOG35-55")) 
saveRDS(vgm_b,"EAE_vgm_b.rds")

#Data architecture and sample counts
names(vgm_allgex[[1]]) #VDJ
names(vgm_allgex[[2]]@meta.data) #GEX
table(vgm_allgex[[1]]$sample_id) #Nr of cells per sample
```

Setting correct sample names and colors for downstream processing and plotting 

```{r, fig.show='hold'}
sample_ids <- c("s1","s2","s3")
short_replace_ids <- c("rMOG_1","rMOG_2", "MOG35-55")

#FOR ALL 3 VGMs generated above
#in VDJ
for (i in 1:3){
  vgm_t[[1]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_t[[1]]$sample_id)
  vgm_b[[1]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_b[[1]]$sample_id)
  vgm_allgex[[1]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_allgex[[1]]$sample_id)
}
vgm_t[[1]]$sample_id <- ordered(as.factor(vgm_t[[1]]$sample_id), levels = short_replace_ids)
vgm_b[[1]]$sample_id <- ordered(as.factor(vgm_b[[1]]$sample_id), levels = short_replace_ids)
vgm_allgex[[1]]$sample_id <- ordered(as.factor(vgm_allgex[[1]]$sample_id), levels = short_replace_ids)

#in GEX
for (i in 1:3){
  vgm_t[[2]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_t[[2]]$sample_id)
  vgm_b[[2]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_b[[2]]$sample_id)
  vgm_allgex[[2]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_allgex[[2]]$sample_id)
}
vgm_t[[2]]$sample_id <- ordered(as.factor(vgm_t[[2]]$sample_id), levels = short_replace_ids)
vgm_b[[2]]$sample_id <- ordered(as.factor(vgm_b[[2]]$sample_id), levels = short_replace_ids)
vgm_allgex[[2]]$sample_id <- ordered(as.factor(vgm_allgex[[2]]$sample_id), levels = short_replace_ids)

#Sample colors were choosen via coolors.co (highly recommended!)
samples_color <- c("#EE8C2F","#F4B67C","#1B9AAA")

#Make a quick dataframe and plot these sample colors for visual reference
short_replace_ids <- ordered(as.factor(short_replace_ids), levels = short_replace_ids)
test_col <- data.frame(pal = samples_color, nam = short_replace_ids , val = runif(3,8,12))

ggplot(test_col, aes(x = nam, y = val, fill= nam)) + geom_bar(stat = "identity") + scale_fill_manual(values = samples_color)

#Group colors rMOG and MOG35-55
groups_color <- c("#F1A155","#1B9AAA")

#also order group_id variable
vgm_t[[1]]$group_id <- ordered(as.factor(vgm_t[[1]]$group_id), levels = c("rMOG", "MOG35-55"))
vgm_b[[1]]$group_id <- ordered(as.factor(vgm_b[[1]]$group_id), levels = c("rMOG", "MOG35-55"))
vgm_allgex[[1]]$group_id <- ordered(as.factor(vgm_allgex[[1]]$group_id), levels = c("rMOG", "MOG35-55"))

vgm_t[[2]]$group_id <- ordered(as.factor(vgm_t[[2]]$group_id), levels = c("rMOG", "MOG35-55"))
vgm_b[[2]]$group_id <- ordered(as.factor(vgm_b[[2]]$group_id), levels = c("rMOG", "MOG35-55"))
vgm_allgex[[2]]$group_id <- ordered(as.factor(vgm_allgex[[2]]$group_id), levels = c("rMOG", "MOG35-55"))

short_replace_ids <- ordered(as.factor(short_replace_ids), levels = short_replace_ids)
test_col <- data.frame(pal = groups_color, nam = short_replace_ids[-2] , val = runif(2,8,12))

ggplot(test_col, aes(x = nam, y = val, fill= nam)) + geom_bar(stat = "identity") + scale_fill_manual(values = groups_color)
```

## GEX overview

We first want an overview of all cells in this dataset and their expression of immune receptors. 
The dataset we are using for this vignette contains both B and T cells (included in the same 10x capture) - we therefore will integrate B cell VDJ info into the vgm_allgex object which already contains T cell VDJ info and use common Seurat plotting functions. 

In the case that either only T cells or only B cells are present, this extra integration step would not be needed.

```{r, fig.show='hold'}
meta_allgex <- vgm_allgex[[2]]@meta.data #getting metadata of allgex object
meta_b <- vgm_b[[2]]@meta.data
for(i in 1:ncol(meta_allgex)){ #converting to character for safety
  meta_allgex[,i] <- as.character(meta_allgex[,i])
  meta_b[,i] <- as.character(meta_b[,i])
}

for(i in 1:nrow(meta_allgex)){ #Going through rows, if a cell barcode (rownames) matches one in the vgm of B cells, this information is copied over
  if(rownames(meta_allgex)[i] %in% rownames(meta_b)){
    meta_allgex[i,] <- meta_b[rownames(meta_b) == rownames(meta_allgex)[i],]
  }
}

vgm_allgex[[2]]@meta.data <- meta_allgex #updating the metadata object
vgm_allgex[[2]]$sample_id <- ordered(as.factor(vgm_allgex[[2]]$sample_id), levels = short_replace_ids)


#Now we add a new column concerning AIR expression, for which we use the VGM celltype column
#By checking the first letter of the VJ_cgene, we make sure that only those cells are labeled for which we do have VDJ information
vgm_allgex[[2]]$AIR <- "VDJ not available"
vgm_allgex[[2]]$AIR[vgm_allgex[[2]]$celltype == "T cell" | substr(vgm_allgex[[2]]$VJ_cgene, 1,1) == "T"] <- "TCR"
vgm_allgex[[2]]$AIR[vgm_allgex[[2]]$celltype == "B cell" | substr(vgm_allgex[[2]]$VJ_cgene, 1,1) == "I"] <- "BCR"

vgm_allgex[[2]]$AIR <- ordered(as.factor(vgm_allgex[[2]]$AIR), levels = c("TCR", "BCR", "VDJ not available"))

#the shuffel = T paramerer in DimPlot is useful for all colored groups to be visible well. As of writing this, there are issues with this and setting split.by. This issue is reported on Github and expected to be fixed. 
plot_out <- Seurat::DimPlot(vgm_allgex[[2]],reduction = "umap", group.by = "AIR", cols = c("brown4","dodgerblue4","grey80"), shuffle = T) + theme(axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position = "top",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold"))  + labs(color = "AIR expression", x = "UMAP1", y = "UMAP2") + guides(color=guide_legend(nrow=1,byrow=TRUE, override.aes = list(size=4))) #adjusting legend rowcount and size of dots

plot_out

#ggsave(plot = plot_out , filename = "allgex_umap_AIR.svg", dpi = 700, width = 6, height = 6.5) 

plot_out <- Seurat::DimPlot(vgm_allgex[[2]],reduction = "umap", group.by = "sample_id", cols = samples_color, shuffle = T, label = F, repel = T, label.size = 10) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Sample", x = "UMAP1", y = "UMAP2")  +guides(color=guide_legend(override.aes = list(size=4)))

plot_out

#ggsave(plot = plot_out, filename = "allgex_umap_sample.svg", dpi = 700, width = 6.5, height = 5)
```

As part of the data overview, we can also explore metrics relating to sequencing quality

```{r, fig.show='hold', warning = FALSE}
#set plot size for all for convenience
wdth <- 4.5
hght <- 6

#cells per sample
metadat <- vgm_allgex[[2]]@meta.data
metadat$sample_id <- ordered(as.factor(metadat$sample_id), levels = short_replace_ids)
metasum <- metadat %>% group_by(sample_id) %>% summarise(n = n())
metasum$sample_id <- ordered(as.factor(metasum$sample_id), levels = short_replace_ids)

plot_out <- ggplot(data = metasum, aes(x = sample_id, y = n, col = as.factor(sample_id), fill = as.factor(sample_id))) + geom_bar(show.legend = F, alpha = 1, stat = "identity", width = 0.6) + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 24),axis.text.y = element_text(size = 24), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=24), legend.position = "none",legend.direction = "vertical", plot.subtitle = element_text(size = 38), plot.title = element_blank()) + labs(title = "", x = "", y = "Recovered cell barcodes", subtitle = "") + scale_fill_manual(values = samples_color) + scale_color_manual(values = samples_color) + geom_text(aes(label = n), size = 11, vjust = -0.5, show.legend = F)+ scale_y_continuous(expand = c(0,0), limits = c(0,8600)) 

plot_out

#ggsave(plot_out, filename = "Cells_per_sample.svg", dpi = 600, width = wdth, height = hght)

#violin number of counts per cell
plot_out <- ggplot(data = metadat, aes(x = sample_id, y = as.numeric(nCount_RNA), col = as.factor(sample_id), fill = as.factor(sample_id))) + geom_violin(show.legend = F, alpha = 1) + geom_jitter(size = 0.8, alpha = 0.3, width = 0.2) + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 24),axis.text.y = element_text(size = 24), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=24), legend.position = "none",legend.direction = "vertical", plot.subtitle = element_text(size = 38), plot.title = element_blank()) + labs(title = "", x = "", y = "Countes per cell", subtitle = "") + scale_fill_manual(values = samples_color) + scale_color_manual(values = samples_color)

plot_out

#ggsave(plot_out, filename = "Counts_per_sample.svg", dpi = 600, width = wdth, height = hght)

#violin number of genes per cell
plot_out <- ggplot(data = metadat, aes(x = sample_id, y = as.numeric(nFeature_RNA), col = as.factor(sample_id), fill = as.factor(sample_id))) + geom_violin(show.legend = F, alpha = 1) + geom_jitter(size = 0.8, alpha = 0.3, width = 0.2) + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 24),axis.text.y = element_text(size = 24), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=24), legend.position = "none",legend.direction = "vertical", plot.subtitle = element_text(size = 38), plot.title = element_blank()) + labs(title = "", x = "", y = "Genes per cell", subtitle = "") + scale_fill_manual(values = samples_color) + scale_color_manual(values = samples_color)

plot_out

#ggsave(plot_out, filename = "genes_per_sample.svg", dpi = 600, width = wdth, height = hght)

#violin % mitochondira 
plot_out <- ggplot(data = metadat, aes(x = sample_id, y = as.numeric(percent.mt), col = as.factor(sample_id), fill = as.factor(sample_id))) + geom_violin(show.legend = F, alpha = 1) + geom_jitter(size = 0.8, alpha = 0.3, width = 0.2) + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 24),axis.text.y = element_text(size = 24), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=24), legend.position = "none",legend.direction = "vertical", plot.subtitle = element_text(size = 38), plot.title = element_blank()) + labs(title = "", x = "", y = "% of MT genes reads", subtitle = "") + scale_fill_manual(values = samples_color) + scale_color_manual(values = samples_color) 
plot_out

#ggsave(plot_out, filename = "MTreads_per_sample.svg", dpi = 600, width = wdth, height = hght)
```

```{r, fig.show='hold', include=FALSE}
rm(list = ls(pattern = "meta"))
rm(vgm_allgex)
gc()
```

## Exploring GEX of T cells by cluster

The following code explores the transcriptional phenotypes of T cells in this dataset.
We will start with some overview UMAPs and work towards cluster classification

```{r, fig.show='hold'}
#By sample 
Seurat::DimPlot(vgm_t[[2]],reduction = "umap", group.by = "sample_id", cols = samples_color, shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Sample", x = "UMAP1", y = "UMAP2") 

#ggsave(plot = last_plot(), filename = "Tcs_umap_sample.svg", dpi = 700, width = 6.5, height = 5)

#By group
Seurat::DimPlot(vgm_t[[2]],reduction = "umap", group.by = "group_id", cols = groups_color, shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Sample", x = "UMAP1", y = "UMAP2") 

#ggsave(plot = last_plot(), filename = "Tcs_umap_group.svg", dpi = 700, width = 6.5, height = 5)

#By cluster
library(pals) #the pals package offers some great varities of palettes, especially for applications were rColorBrewer palettes contain too few distinct colors

Seurat::DimPlot(vgm_t[[2]],reduction = "umap", group.by = "seurat_clusters", cols = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)], shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Cluster", x = "UMAP1", y = "UMAP2") 

#ggsave(plot = last_plot(), filename = "Tcs_umap_clusters.svg", dpi = 700, width = 6.5, height = 5)
```

We see clear overlap between samples. Almost all clusters are composed of cells of all samples and groups. We can visualize proportions using stacked barplots.

```{r, fig.show='hold'}
cl_members <- GEX_proportions_barplot(GEX = vgm_t[[2]],source.group = "sample_id", target.group = "seurat_clusters", stacked.plot = T)+ scale_fill_manual(values = samples_color, labels = short_replace_ids) + theme(panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "bottom", plot.title = element_blank(), legend.title = element_text(size = 30, face = "bold")) + labs(title = "", x = "", y = "% of cells of cluster", fill = "Sample") + scale_y_continuous(expand = c(0,0))

cl_members 

#ggsave(last_plot(), filename = "Tc_cluster_membership_sample_new.svg", dpi = 400, width = 10, height = 8)
```

Now we can start investigating what the clusters actually represent:
Step 1: Project common markers to get a basic idea

```{r, fig.show='hold'}
library(scales)

FeaturePlot(vgm_t[[2]], features = c("CD3G","CD4","CD8A","FOXP3"), reduction = "umap", label = F, repel = F, pt.size = 1.3) & theme(axis.text.y = element_blank(),axis.text.x = element_blank(),axis.line=element_line(size = 1),axis.ticks = element_blank(), legend.position = c(0.5,0.1), legend.direction = "horizontal", axis.title = element_text(), plot.title = element_text(size = 27, face = "bold")) & labs(x = "UMAP1", y = "UMAP2") & scale_color_gradient2(low = "grey80", mid = "grey80", high = "#DD513AFF", oob = squish_infinite) #We position the legend inside the plot using legend.position, to save some space. Further we use the "squish_infinite" function of the scales package to easily reppresent outlier values without compromising overall color scales.  

#ggsave(plot = last_plot(), filename = "Tcs_umap_Tcell_basic.svg", dpi = 400, width = 10, height = 10)
```

For a more in depth look at marker expression, we use the dottile function. 

```{r, fig.show='hold'}
#Check what genes are available with this searchable dataframe view (at least in R studio)
#View(as.data.frame(rownames(vgm_t)))

genes_to_plot <- c("CD4","CD8A","CD28","ICOS","CD40LG","CTLA4","PDCD1","LAG3","S1PR1","PTPRC","SELL","TBX21","GATA3","SPI1","IRF4","BCL6","STAT4","STAT6","FOXP3","IKZF2", "MKI67","TCF7","TOX", "GZMB","TNF","IFNG","TGFB1","IL10","IL17A","CSF2","IL2RA","IL4RA","IL12RB1","CXCR3","CCR5","CCR8", "SOX4","RORA","RORC","RUNX3","HIF1A")

linesh <- c(1:length(genes_to_plot)) + 0.5 #these lines are for easier viewing of the plot
plot_out <- GEX_dottile_plot(GEX = vgm_t[[2]], genes = genes_to_plot, group.by = "seurat_clusters", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 0, hjust = 0.5,vjust = 0.5)) +scale_size_binned(range = c(0.01,12)) + geom_hline(yintercept = linesh, col = "black", size = 0.9)+ scale_color_viridis_c(option = "B", end = 0.85) #we end the viridis palette at .85 specifically, to avoid bright yellow spots that can be hard to view when printed

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Tc_dottile.svg", dpi = 400, width = 10, height = 15)
```

We can now find cluster specific genes by DEG analysis. 
Here we use the GEX_cluster_genes to find genes that are deferentially expressed in one cluster relative to all other cells. This function is based on the Seurat::FindAllMarkers function but allows the user to filter out mitochondrial and ribosomal proteins from the list and can further be used as input to other plotting functions. 

```{r, fig.show='hold'}
## Warning: running this function will take a while
#gene_expression_cluster <- GEX_cluster_genes(vgm_t[[2]],min.pct = 0.25) 
#saveRDS(gene_expression_cluster, file = "Tcs_gene_expression_clusters.rds")

gene_expression_cluster <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/Tcs_gene_expression_clusters.rds")
#Recommended to get an overview of the output format
names(gene_expression_cluster[[1]])
#View(gene_expression_cluster[[1]])
```

We can visualize these cluster defining genes both as a heatmap, a dotmap, and as volcano plots

```{r, fig.show='hold'}
#heatmap
EAE_heatmap_clusters <- GEX_cluster_genes_heatmap(GEX = vgm_t[[2]], GEX_cluster_genes.output = gene_expression_cluster, n.genes.per.cluster = 5,max.cell = 30, metric = "top_logFC", group.colors = kelly(n = 17)[c(2,3,17,7:13,6,16)]) + scale_fill_gradientn(colors = c("white","grey80","darkorchid3"), oob = squish_infinite) + theme(axis.text.y = element_text(size = 13),axis.text.x = element_blank(), legend.position = "none") #here we picked top genes by P value. We recommend to also check top genes by p value for consistency

EAE_heatmap_clusters

#ggsave(plot = last_plot(), filename = "Tc_cluster_heatmap_top5_logFC.svg", dpi = 400, width = 10, height = 10)
```

The dottile function can be run directly from the gene selection that made by the heatmap function

```{r, fig.show='hold'}
cldef <- EAE_heatmap_clusters$data #getting the source data table of the heatmap from above
length(unique(cldef$Feature)) #how many genes were selected

genes_to_plot <- as.character((unique(cldef$Feature)))
lines <- c(seq(0,58,5)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_t[[2]], genes = rev(genes_to_plot), group.by = "seurat_clusters", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 0, hjust = 0.5,vjust = 0.5))  +scale_size_binned(range = c(0.01,12)) + geom_hline(yintercept = lines, col = "black", size = 0.9) + guides(color = guide_legend(reverse=T))+ scale_color_viridis_c(option = "B", end = 0.85)

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Tc_cluster_dottile_top5_logFC.svg", dpi = 400, width = 10, height = 22)
```

And some more DEG analysis between clusters: 
What are differences between cluster 0 and 1.

```{r, fig.show='hold', eval = FALSE}
library(ggrepel)
#Here we return a volcano plot where the top 40 genes are labelled by geom_text_repel. Further the genes CD28 and ICOS are labelled independently of their p.value
plot_out <- GEX_DEgenes(GEX = vgm_t[[2]], grouping.column = "seurat_clusters", group1 = "0", group2 = "1", return.plot = "volcano", logFC = FALSE, label.n.top.genes = 40, genes.to.label = c("CD28", "ICOS"))

#plot_out[[1]] #For source table
plot_out[[2]] #For plot
```

With additional ggplot coding we can produce summary volcano plots with the defining genes of each cluster. We use the gtable package to recolor the facet wrap bars to match cluster colors 

```{r, fig.show='hold'}
#extracting top 15 genes per cluster by fold change
top_genes_per_cluster <- GEX_topN_DE_genes_per_cluster(GEX_cluster_genes.output = gene_expression_cluster, n.genes = 15, by_FC = T)

#for the later recoloring of bars in the facet wrap plot, we need an even number of facets in the plot i.e. clusters
#Here we add an empty one to match this
extr_cluster <- subset(top_genes_per_cluster, cluster == "0")
extr_cluster$SYMBOL <- "none"
extr_cluster$cluster <- 11
top_genes_per_cluster <- rbind(top_genes_per_cluster, extr_cluster)

#Some of these genes have an adj p value of 0, which results in strange positions when plotting. Here we set them to an low but non-0 value
top_genes_per_cluster$p_val_adj[top_genes_per_cluster$p_val_adj == 0] <- 1e-299

#generating a large facet plot 
plot_out <- ggplot(top_genes_per_cluster, aes(x = avg_logFC, y = pct.1*100, col = -log10(p_val_adj))) + geom_point(show.legend = T, size = 3, alpha = 1)  + theme(panel.background = element_blank(), axis.title = element_text(size = 30),axis.text = element_text(size = 21),plot.title = element_blank(), axis.line = element_line(size = 2), strip.text = element_text(size = 30), axis.ticks = element_line(color = "black", size = 2), axis.ticks.length = unit(0.25,"cm"), legend.key = element_rect(colour = "white"), legend.key.width = unit(1.2, "cm"), strip.background = element_rect(colour = "white", fill = "white"), strip.text.x = element_text(color = "white", face = "bold"), legend.position = "top", legend.direction = "horizontal", legend.title = element_text(size = 28, face = "bold"), legend.text = element_text(size = 21)) + labs(x = "log2(FC)", y = "% of expressing cells in cluster", col = "-log10(adj p value)") + geom_text_repel(aes(label = SYMBOL), size = 6, nudge_x = 0, nudge_y = 0, segment.alpha = 1, max.overlaps = 14) +scale_color_viridis_c(option = "B", end = 0.85) + facet_wrap(~cluster,ncol = 6,scales = "free") + geom_vline(xintercept = 0, linetype = 2, size = 1.5, alpha = 0.6) 

#recoloring the facet bars (code snippet from Stackoverflow) !!! For this to work, the facet wrap must contain only full rows (i.e. in this case with two rows, plot numbers must be even) !!!
library(gtable)

cols <- c(pals::kelly(n = 17)[c(10,11,12,13,6,16,2,3,17,7,8,9)]) #colors
g <- ggplot_gtable(ggplot_build(plot_out))
stripr <- which(grepl('strip', g$layout$name))
stripr
fills <- cols
k <- 1
for (i in stripr) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}

plot(g) 

#ggsave(g, filename = "Tcs_degbycluster_wrap.svg", dpi = 700, width = 24, height = 14.5)
g <- NULL
```

Now to examine coexpression of selected genes across different populations of T cells:

```{r, fig.show='hold'}
#Cluster 0 => show association of active memory leaning phenotype with Tgfb1 expression

genes_to_plot <- c("CD4","CD28","S1PR1","PTPRC","IL2RA","TBX21","GATA3","STAT4","TCF7","NKG7","TGFB1","IFNG")

cur_c <- subset(vgm_t[[2]], seurat_clusters == "0")

plot_out <- GEX_coexpression_coefficient(GEX = cur_c, genes = genes_to_plot, subsample.n = "none") + theme( axis.title.x = element_text(), legend.position = "right")  +scale_size_binned(range = c(0.01,18)) +scale_color_viridis_c(option = "B", end = 0.85) + labs(color = "% double \n  positive", size = "% positive", title = "Cluster 0")

plot_out

#ggsave(plot_out, filename = paste0("Tcs_selected_coex_cluster_0.svg"), dpi = 400, width = 10.7, height = 9)

#Cluster 8 Examine FOXP3 Tregs

genes_to_plot <- c("CD4","CD28","PTPRC","CTLA4","IL2RA","FOXP3","TOX","IKZF2","GZMB","IL10","TGFB1","IFNG")

cur_c <- subset(vgm_t[[2]], seurat_clusters == "8")
  
plot_out <- GEX_coexpression_coefficient(GEX = cur_c, genes = genes_to_plot, subsample.n = "none") + theme(axis.title.x = element_text(), legend.position = "right")  +scale_size_binned(range = c(0.01,18), n.breaks = 4) +scale_color_viridis_c(option = "B", end = 0.85) + labs(color = "% double \n positive", size = "% positive", title = "Cluster 8")

plot_out

#ggsave(plot_out, filename = paste0("Tcs_selected_coex_cluster_8.svg"), dpi = 400, width = 10.7, height = 9)
```

We also explore coexpression for specifc gene pairs

```{r, fig.show='hold'}
#FIRST CD40LG vs. BCL6 to examine the presence of Tfh cells
plot_out <-  GEX_scatter_coexpression(GEX = vgm_t[[2]], gene.1 = "CD40LG", gene.2 = "BCL6", color.theme = "dodgerblue4")

#ggsave(plot = plot_out, filename = "Tc_scatter_CD40LGBCL6dp.svg", dpi = 400, width = 10, height = 10)

#Using the GEX_DE_genes function we can view the full transcriptional profile of such double positive cells 
#add a column
vgm_t[[2]]$Tfh <- FALSE
tfh_markers <- FetchData(vgm_t[[2]], vars = c("CD40LG", "BCL6"))
vgm_t[[2]]$Tfh[tfh_markers$CD40LG > 0 & tfh_markers$BCL6 > 0] <- TRUE

plot_out <- GEX_DEgenes(GEX = vgm_t[[2]], grouping.column = "Tfh", group1 = "TRUE", group2 = "FALSE", return.plot = "volcano", logFC = FALSE, label.n.top.genes = 15, genes.to.label = c("CD28", "ICOS"))

plot_out[[2]]

#Also validating IFNG & TGFB1 coexpression
 
plot_out <-  GEX_scatter_coexpression(GEX = vgm_t[[2]], gene.1 = "IFNG", gene.2 = "TGFB1", color.theme = "dodgerblue4")

#ggsave(plot = plot_out, filename = "Tc_scatter_IFNGTGFB1dp.svg", dpi = 400, width = 10, height = 10)
```

## Comparing GEX of T cells by group

A key goal of this study is the comparison between T cells in EAE initiated by two different autoantigens: rMOG and MOG35-55
Here we explicitly compare the T cells from the two conditions.

```{r, fig.show='hold'}
#first a membership barplot to check if T cells from either immunizations cluster differently
cl_members <- GEX_proportions_barplot(GEX = vgm_t[[2]],source.group = "group_id", target.group = "seurat_clusters", stacked.plot = T)+ scale_fill_manual(values = groups_color, labels = c("rMOG","MOG35-55")) + theme(panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "bottom", plot.title = element_blank(), legend.title = element_text(size = 30, face = "bold")) + labs(title = "", x = "", y = "% of cells of cluster", fill = "Sample") + scale_y_continuous(expand = c(0,0))

cl_members 

#ggsave(last_plot(), filename = "Tc_cluster_membership_group.svg", dpi = 400, width = 10, height = 8)

#Get DEGs 
exp_DEGs <- GEX_DEgenes(GEX = vgm_t[[2]], min.pct = 0.25, grouping.column = "group_id", group1 = "rMOG", group2 = "MOG35-55", return.plot = "none", logFC = F) #here we separate the function call from the ggplot for more flexibility. output[[1]] of the GEX_DEgenes is a dataframe with DEGs

plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 1)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "rMOG vs. MOG33-55") + geom_text_repel(data = subset(exp_DEGs[[1]], p_val_adj < 0.00001), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 20)+scale_color_gradient2(low = samples_color[3], mid = "grey60", high = samples_color[1], midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_volc_bygroup.svg", dpi = 400, width = 13, height = 11)

#DEGs after p correction
length(exp_DEGs[[1]]$p_val_adj[exp_DEGs[[1]]$p_val_adj < 0.05])
```

For some more insight into these genes we run Gene ontology analysis
This outputs plots as PDFs, which are not displayed in this vignette

```{r, fig.show='hold'}
#ontology_EAE <- GEX_GOterm(GEX.cluster.genes.output = list(exp_DEGs[[1]]), topNgenes = 40, go.plots = T)
```

Given that cluster 4 and 7 are almost exclusive to rMOG, we test whether the DEGs calculated above remain stable even if these two clusters are excluded

```{r, fig.show='hold'}
#Subsetting the GEX object
cells_to_keep <- colnames(vgm_t[[2]])[!vgm_t[[2]]$seurat_clusters %in% c(7,4)]
for_deg <- subset(vgm_t[[2]], cells = cells_to_keep) #subset using Seurat parameters

#plot umap to verify that exclusion was successfull
Seurat::DimPlot(for_deg,reduction = "umap", group.by = "seurat_clusters", cols = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)], shuffle = T) + theme(axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Cluster", x = "UMAP1", y = "UMAP2") 

exp_DEGs <- GEX_DEgenes(GEX = for_deg, min.pct = 0.25, grouping.column = "group_id", group1 = "rMOG", group2 = "MOG35-55", return.plot = "none", logFC = F) #here we separate the function call from the ggplot for more flexibility. output[[1]] of the GEX_DEgenes is a dataframe with DEGs

plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 1)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "rMOG vs. MOG33-55 ex cluster 4 & 7") + geom_text_repel(data = subset(exp_DEGs[[1]], p_val_adj < 0.00001), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 20)+scale_color_gradient2(low = samples_color[3], mid = "grey60", high = samples_color[1], midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_volc_bysample_excl47.svg", dpi = 400, width = 10, height = 15)
```

For completeness we also plot a dotmap of selected genes across groups

```{r, fig.show='hold'}
genes_to_plot <- c("CD4","CD8A","CD69","CD28","PDCD1","LAG3","S1PR1","IL2RA","FOXP3","TCF7","TOX", "GZMB","IFNG","TGFB1")

linesh <- c(1:length(genes_to_plot)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_t[[2]], genes = genes_to_plot, group.by = "group_id", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 45, hjust = 1,vjust = 1), axis.title.x.bottom = element_blank(), axis.ticks = element_line(size = 1), axis.ticks.length = unit(0.15,"cm"))  +scale_size_binned(range = c(0.1,12), n.breaks = 5)  + geom_hline(yintercept = linesh, col = "black", size = 0.9) + scale_color_viridis_c(option = "B", end = 0.85)

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Tc_dottile_group.svg", dpi = 400, width = 5.5, height = 7)
```

## T cells integration with aged CNS dataset

As part of an earlier study, our group examined phenotypes of T cells in the CNS of aged mice. They found increased numbers of lymphocytes and increased clonal expansion, especially of CD8+ T cells in aged vs. young animals. 

(Yermanos, A. et al. Single-cell immune repertoire and transcriptome sequencing reveals that clonally expanded and transcriptionally distinct lymphocytes populate the aged central nervous system in mice. Proc. R. Soc. B Biol. Sci. 288, (2021).)

We wondered how different cells of EAE and aged CNS are transcriptionally. 

First we run a new VGM to integrate these datasets. While we show the case where the data is on the user's local PC, please check out the PlatypusDB vignette to see how this could have been done directly from the database.

```{r, fig.show='hold'}
VDJ.out.directory.list_T <- list()
VDJ.out.directory.list_T[[1]] <- c("~/EAE/tcell/S1") #rMOG_1 
VDJ.out.directory.list_T[[2]] <- c("~/EAE/tcell/S1") #rMOG_2
VDJ.out.directory.list_T[[3]] <- c("~/EAE/tcell/S1") #MOG35-55
VDJ.out.directory.list_T[[4]] <- c("~/Aged_CNS/vdj/S1") # 3 Months pool
VDJ.out.directory.list_T[[5]] <- c("~/Aged_CNS/vdj/S2") # 12 months pool
VDJ.out.directory.list_T[[6]] <- c("~/Aged_CNS/vdj/S3") # 18 months pool
VDJ.out.directory.list_T[[7]] <- c("~/Aged_CNS/vdj/S4") # 18 months alone

GEX.out.directory.list <- list()
GEX.out.directory.list[[1]] <- c("~/EAE/gex/S1") #rMOG_1
GEX.out.directory.list[[2]] <- c("~/EAE/gex/S1") #rMOG_2
GEX.out.directory.list[[3]] <- c("~/EAE/gex/S1") #MOG35-55
GEX.out.directory.list[[4]] <- c("~/Aged_CNS/gex/S1") #3 months old pooled
GEX.out.directory.list[[5]] <- c("~/Aged_CNS/gex/S2") #12 months old pooled
GEX.out.directory.list[[6]] <- c("~/Aged_CNS/gex/S3") #18 months old pooled
GEX.out.directory.list[[7]] <- c("~/Aged_CNS/gex/S4") #18 months old single
```

```{r, include = F}
VDJ.out.directory.list_T <- list()
#EAE
VDJ.out.directory.list_T[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/tcell/S1") #Protein acute brain + spinal pooled
VDJ.out.directory.list_T[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/tcell/S2") #Protein acute brain + spinal pooled
VDJ.out.directory.list_T[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/tcell/S3") #Peptide acute brain + spinal
VDJ.out.directory.list_T[[4]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S1") #3 months old pooled
VDJ.out.directory.list_T[[5]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S2") #12 months old pooled
VDJ.out.directory.list_T[[6]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S3") #18 months old pooled
VDJ.out.directory.list_T[[7]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S4") #18 months old single

GEX.out.directory.list <- list()
#EAE
GEX.out.directory.list[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S1") #Protein acute brain + spinal pooled
GEX.out.directory.list[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S2") #Protein acute brain + spinal pooled
GEX.out.directory.list[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S3") #Peptide acute brain + spinal
GEX.out.directory.list[[4]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S1") #3 months old pooled
GEX.out.directory.list[[5]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S2") #12 months old pooled
GEX.out.directory.list[[6]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S3") #18 months old pooled
GEX.out.directory.list[[7]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S4") #18 months old single

gc()

vgm_EAEaged <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_EAEaged.rds")

```

Run VGM including both the EAE and the aged CNS dataset

```{r, fig.show='hold'}

#vgm_EAEaged <- VDJ_GEX_matrix(VDJ.out.directory.list = VDJ.out.directory.list_T,
#                             GEX.out.directory.list = GEX.out.directory.list,
#                             exclude.on.cell.state.markers = c("CD19+", "CD3E-", "EBF1+"), 
#                             group.id = c("EAE", "EAE", "EAE", "aged CNS", "aged CNS", "aged CNS", "aged CNS"), get.VDJ.stats = F) 

#saveRDS(vgm_EAEaged,"EAE_vgm_EAEaged.rds")



sample_ids <- c("s1","s2","s3","s4","s5","s6", "s7")
short_replace_ids <- c("rMOG_1", "rMOG_2","MOG35-55", "3m", "12m", "18m", "18m single")

for (i in 1:7){
  vgm_EAEaged[[2]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_EAEaged[[2]]$sample_id)
}
vgm_EAEaged[[2]]$sample_id <- ordered(as.factor(vgm_EAEaged[[2]]$sample_id), levels = unique(short_replace_ids))

unique(vgm_EAEaged[[2]]$sample_id)

samples_color_aged <- c(samples_color,"#71716F", "#577590", "#B2B7A4", "#D8DAD3")

short_replace_ids <- ordered(as.factor(short_replace_ids), levels = unique(short_replace_ids))
test_col <- data.frame(pal = samples_color_aged, nam = unique(short_replace_ids) , val = runif(7,8,12))

#Color test ggplot
ggplot(test_col, aes(x = nam, y = val, fill= nam)) + geom_bar(stat = "identity") + scale_fill_manual(values = samples_color_aged)

#Nr. cells per sample 
table(vgm_EAEaged[[2]]$sample_id)
```

Next: General UMAP to show transcriptional landscape

```{r, fig.show='hold'}
Seurat::DimPlot(vgm_EAEaged[[2]],reduction = "umap", group.by = "sample_id", cols = samples_color_aged, shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold"), legend.text = element_text(size = 15)) + labs(color = "Sample", x = "UMAP1", y = "UMAP2") + guides(colour = guide_legend(override.aes = list(size=5)))

#ggsave(plot = last_plot(), filename = "Tcs_eaeaged_umap_sample.svg", dpi = 700, width = 6.5, height = 5)

Seurat::DimPlot(vgm_EAEaged[[2]],reduction = "umap", group.by = "seurat_clusters", cols = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)], shuffle = T, label =T, label.size = 7 ) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold"), legend.text = element_text(size = 15)) + labs(color = "Cluster", x = "UMAP1", y = "UMAP2")+ guides(colour = guide_legend(override.aes = list(size=5)))

#ggsave(plot = last_plot(), filename = "Tcs_eaeaged_umap_cluster.svg", dpi = 700, width = 6.5, height = 5)
```

Next we visualize cluster membership with a barplot

```{r, fig.show='hold'}
cl_members <- GEX_proportions_barplot(GEX = vgm_EAEaged[[2]],source.group = "sample_id", target.group = "seurat_clusters", stacked.plot = T)+ scale_fill_manual(values = samples_color_aged) + theme(plot.margin = margin(t = 0.7,r = 2.5, unit = "cm"),panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 1), axis.ticks = element_line(size = 1), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "none", plot.title = element_blank(), legend.title = element_text(size = 25, face = "bold"), legend.text = element_text(size = 19)) + labs(title = "", x = "", y = "% of cells of cluster", fill = "Sample") + scale_y_continuous(expand = c(0,0))
cl_members 

#ggsave(last_plot(), filename = "Tcs_eaeaged_Cluster_membership.svg", dpi = 400, width = 7.5, height = 5)
```

Dotmaps again come in useful for many genes and conditions

```{r, fig.show='hold'}
genes_to_plot <- c("CD4","CD8A","CD69","CD28","ICOS","CD40LG","PDCD1","LAG3","S1PR1","PTPRC","SELL","IL7R","STAT4","FOXP3", "MKI67","TCF7","TOX", "GZMB","GZMK","IFNG","TGFB1","IL17A","CXCR3","CCR5","ITGA4","ITGB2")

linesh <- c(1:length(genes_to_plot)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_EAEaged[[2]], genes = genes_to_plot, group.by = "sample_id", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 45, hjust = 1,vjust = 1), axis.title.x.bottom = element_blank(), axis.ticks = element_line(size = 1), axis.ticks.length = unit(0.15,"cm"))  +scale_size_binned(range = c(0.1,12), n.breaks = 5)  + geom_hline(yintercept = linesh, col = "black", size = 0.9) + scale_color_viridis_c(option = "B", end = 0.85)

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Tc_eaeaged_dottile.svg", dpi = 400, width = 8, height = 11)
```

Differentially expressed genes by condition: 
Given the higher number of CD8+ T cells in aged CNS, we can also split these DEG calculations by T cell subtype

```{r, fig.show='hold'}
vgm_EAEaged[[2]]$CD4_ <- "neg"
vgm_EAEaged[[2]]$CD4_dat <- FetchData(vgm_EAEaged[[2]], "CD4")
vgm_EAEaged[[2]]$CD4_[which(vgm_EAEaged[[2]]$CD4_dat > 0)] <- "CD4"
vgm_EAEaged[[2]]$CD4_ <- paste0(vgm_EAEaged[[2]]$group_id, " ", vgm_EAEaged[[2]]$CD4_)

table(vgm_EAEaged[[2]]$CD4_)

sample_DEGs <- GEX_DEgenes(vgm_EAEaged[[2]], min.pct = 0.25, grouping.column = "CD4_", group1 = "EAE CD4", group2 = "aged CNS CD4", return.plot = "none", logFC = F)

#plot volcano
plot_out <- ggplot(sample_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 1), axis.ticks = element_line(size = 1, color = "black"), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "CD4+ EAE vs. Aged") + geom_text_repel(data = subset(sample_DEGs[[1]], -log10(p_val_adj) > 20 | SYMBOL %in% c("TCF7", "GZMB", "CD28", "CD69", "HIF1A", "TOX","GZMK")), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 40)+scale_color_gradient2(low = samples_color_aged[4], mid = "grey60", high = samples_color_aged[2], midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_eaeaged_CD4volc.svg", dpi = 400, width = 8, height = 9.2)

#GIVEN the high number of CD8 in Aged: detailed comparison of CD8+ Tcs 

vgm_EAEaged[[2]]$CD8_ <- "neg"
vgm_EAEaged[[2]]$CD8_dat <- FetchData(vgm_EAEaged[[2]], "CD8A")
vgm_EAEaged[[2]]$CD8_[which(vgm_EAEaged[[2]]$CD8_dat > 0)] <- "CD8"
vgm_EAEaged[[2]]$CD8_ <- paste0(vgm_EAEaged[[2]]$group_id, " ", vgm_EAEaged[[2]]$CD8_)

vgm_EAEaged[[2]]$sample_id <- vgm_EAEaged[[2]]$CD8_

sample_DEGs <- GEX_DEgenes(vgm_EAEaged[[2]], min.pct = 0.25, grouping.column = "CD8_", group1 = "EAE CD8", group2 = "aged CNS CD8", return.plot = "none", logFC = F)

#plot volcano
plot_out <- ggplot(sample_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 1), axis.ticks = element_line(size = 1, color = "black"), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "CD8+ EAE vs. Aged") + geom_text_repel(data = subset(sample_DEGs[[1]], -log10(p_val_adj) > 37 | SYMBOL %in% c("TCF7", "GZMB", "CD28", "CD69", "HIF1A", "TOX","GZMK")), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 40)+scale_color_gradient2(low = samples_color_aged[4], mid = "grey60", high = samples_color_aged[2], midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_eaeaged_new_CD8volc.svg", dpi = 400, width = 8, height = 9.2)
```

## Exploring VDJ of T cells

After having explored transcriptional phenotypes of T cells we now turn to their TCR repertoires. 
We can plot circular expansion graphs

```{r, fig.show='hold'}
#with VGM counts
plots_out <- VDJ_clonal_donut(vgm_t[[1]], counts.to.use = "VGM", expanded.colors = c("#68228B","#BF3EFF","#E066FF","#8968CD"), non.expanded.color = "#000000", label.size = 9, total.label.vjust = 2) #Middle label position slightly adjusted to fit. Please only use these parameters after having set on a final output format for the plot
#here we use counts of clones from the VGM (i.e. the nr. of rows/cells of every clone) This can also be set to "10x" to use counts from the clonotype_frequency column based on original Cellranger output. The two may diverge if a large number of overlapping barcodes were filtered out or if the clonotyping strategy was changed 

for(i in 1:length(plots_out)){ #saving in a loop
  print(plots_out[[i]])
  #ggsave(plots_out[[i]], filename = paste0("Donut_VGMcounts",plots_out[[i]]$labels$title ,".png"), dpi = 400, width = 5, height = 5, bg = "transparent")
}


#check mean expanded clone frequency by group for 10x counts
vgm_t[[1]]$clonotype_frequency <- as.numeric(vgm_t[[1]]$clonotype_frequency) #making sure 
vgm_t[[1]] %>% group_by(sample_id) %>% summarize(mean_cl_freq = mean(clonotype_frequency), median_cl_freq = median(clonotype_frequency))
```

We continue by calculating common diversity metrics
Here we use VDJ and VJ CDR3s seqs as a basis

```{r, fig.show='hold'}
#NO SUBSAMPLING
out_list <- list() #loop to compile results for all metrics
for(i in c("richness","bergerparker","ginisimpson","shannonevenness")){
  out <- VDJ_diversity(VDJ = vgm_t[[1]],feature.columns = c("VDJ_cdr3s_aa","VJ_cdr3s_aa"),grouping.column = "sample_id", metric = i, platypus.version = "v3", subsample.to.same.n = F)$data
  out$metric_used <- i
  out_list[[i]] <- out
}
out <- bind_rows(out_list) #generating a common dataframe

#lets plot this
out$index <- ordered(as.factor(out$index), levels = c("Species richness", "Berger-Parker index", "Gini-Simpson index", "Shannon evenness"))
plot_out <- ggplot(out, aes(x = groups, y = metric, fill = groups))  + labs(title = "", x = "", y = "") + theme(panel.background = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") + geom_bar(width = 0.6,show.legend = F, stat = "identity") + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 30),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), strip.background = element_rect(fill = "white", color = "white")) + scale_y_continuous(expand = c(0,0))+ scale_fill_manual(values= samples_color)+ facet_wrap(~index, scales = "free") 

plot_out

#ggsave(plot_out, filename = "Tcs_rep_diversity_nosubsampling.svg", dpi = 600, width = 10, height = 8)

#SUBSAMPLING
out_list <- list()
for(i in c("richness","bergerparker","ginisimpson", "shannonevenness")){
  out <- VDJ_diversity(VDJ = vgm_t[[1]],feature.columns = c("VDJ_cdr3s_aa","VJ_cdr3s_aa"),grouping.column = "sample_id",metric = i, platypus.version = "v3", subsample.to.same.n = T)$data #subsampling down to the smallest repertoire
  out$metric_used <- i
  out_list[[i]] <- out
}
out <- bind_rows(out_list)

#lets plot this too
out$index <- ordered(as.factor(out$index), levels = c("Species richness", "Berger-Parker index", "Gini-Simpson index", "Shannon evenness"))
plot_out <- ggplot(out, aes(x = groups, y = metric, fill = groups))  + labs(title = "", x = "", y = "") + theme(panel.background = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") + geom_bar(width = 0.6,show.legend = F, stat = "identity") + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 30),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), strip.background = element_rect(fill = "white", color = "white")) + scale_y_continuous(expand = c(0,0))+ scale_fill_manual(values= samples_color)+ facet_wrap(~index, ncol = 2, scales = "free") 

plot_out

#ggsave(plot_out, filename = "Tcs_rep_diversity_subsampled.svg", dpi = 600, width = 8, height = 10)
```

V gene usage is an important indicator of repertoire similarity or dissimilarity. 
1. stacked barplot 

```{r, fig.show = 'hold'}
levels(vgm_t[[1]]$sample_id)

vgm_t[[1]]$VDJ_vgene <- gsub("TRBV12-2\\+TRBV13-2", "TRBV13-2",vgm_t[[1]]$VDJ_vgene) #A single expanded clone was noted to have 2 heavy chains with 2 different V genes. Here we remove the infrequent of the two.

EAE_Vgene_usage_barplot_b <- VDJ_Vgene_usage_stacked_barplot(vgm_t[[1]], HC.gene.number = 12, LC.Vgene = F, LC.gene.number = 12, platypus.version = "v3")

#ggplot magic
plot_out <- EAE_Vgene_usage_barplot_b + theme(panel.background = element_blank(),axis.text = element_text(size = 25), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=28),legend.position = "right",axis.text.x = element_text(angle = 30,vjust = 1, hjust=1), plot.title = element_blank(), axis.title.x = element_blank(), plot.margin = margin(t = 0.7, unit = "cm"), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),panel.border = element_blank(), legend.title = element_text(face = "bold"))+ scale_fill_manual(values = pals::stepped2(20))+ scale_x_discrete(breaks=c("1","2","3"),labels=c("rMOG_1", "rMOG_2", "MOG35-55")) + labs(y = "relative usage", fill = "VDJ Vgene")

plot_out

#ggsave(plot = plot_out, filename = paste0("Tcs_VDJ_vgene_by_sample.svg"), dpi = 400, width = 8, height = 7)

#VJ
EAE_Vgene_usage_barplot_b <- VDJ_Vgene_usage_stacked_barplot(vgm_t[[1]], HC.gene.number = 12, LC.Vgene = T, LC.gene.number = 12, platypus.version = "v3")

plot_out <- EAE_Vgene_usage_barplot_b + theme(panel.background = element_blank(),axis.text = element_text(size = 25), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=28),legend.position = "right",axis.text.x = element_text(angle = 30,vjust = 1, hjust=1), plot.title = element_blank(), axis.title.x = element_blank(), plot.margin = margin(t = 0.7, unit = "cm"), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),panel.border = element_blank(), legend.title = element_text(face = "bold"))+ scale_fill_manual(values = pals::stepped2(20))+ scale_x_discrete(breaks=c("1","2", "3"),labels=c("rMOG_1", "rMOG_2", "MOG35-55")) + labs(y = "relative usage", fill = "VJ Vgene") 

plot_out

#ggsave(plot = plot_out, filename = paste0("Tcs_VJ_vgene_by_sample.svg"), dpi = 400, width = 8, height = 7)
```

2. Using Circos plots 
These will generate PDF files, which are not contained within this vignette
```{r, fig.show = 'hold', eval = FALSE}
pdf(file = "Tcs_circos_rMOG_1.pdf", onefile = T, width = 6, height = 6)
for_cir <- subset(vgm_t[[1]], sample_id == "rMOG_1")
for_cir$sample_id <- as.character(for_cir$sample_id)
circos <- VDJ_VJ_usage_circos(VDJ = for_cir, A.or.B = "both", label.threshold = 5, cell.level = T, clonotype.per.gene.threshold = 20, c.count = T, platypus.version = "v3", filter1H1L = T)
dev.off()

pdf(file = "Tcs_circos_rMOG_2.pdf", onefile = T, width = 6, height = 6)
for_cir <- subset(vgm_t[[1]], sample_id == "rMOG_2")
for_cir$sample_id <- as.character(for_cir$sample_id)
circos <- VDJ_VJ_usage_circos(VDJ = for_cir, A.or.B = "both", label.threshold = 5, cell.level = T, clonotype.per.gene.threshold = 20, c.count = T, platypus.version = "v3", filter1H1L = T)
dev.off()

pdf(file = "Tcs_circos_MOG3555.pdf", onefile = T, width = 6, height = 6)
for_cir <- subset(vgm_t[[1]], sample_id == "MOG35-55")
for_cir$sample_id <- as.character(for_cir$sample_id)
circos <- VDJ_VJ_usage_circos(VDJ = for_cir, A.or.B = "both", label.threshold = 5, cell.level = T, clonotype.per.gene.threshold = 10, c.count = T, platypus.version = "v3", filter1H1L = T)
dev.off()
```

Public clones are often of great interest to assess clonal convergence. The following VDJ_overlap_heatmap function quantifies this for the vgm object.

```{r, fig.show='hold'}
#using the unfiltered (VDJ comb here. All cells, also those with less or more than 1VDJ 1VJ are included)
VDJv_overlap <- VDJ_overlap_heatmap(vgm_t[[1]],feature.columns = c("VDJ_vgene") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3", add.barcode.table = T) 
length(unique(vgm_t[[1]]$VDJ_vgene))

VJv_overlap <- VDJ_overlap_heatmap(vgm_t[[1]],feature.columns = c("VJ_vgene") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3", add.barcode.table = T)
length(unique(vgm_t[[1]]$VJ_vgene))

VJ_overlap <- VDJ_overlap_heatmap(vgm_t[[1]],feature.columns = c("VJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3")
length(unique(vgm_t[[1]]$VJ_cdr3s_aa))

VDJ_overlap <- VDJ_overlap_heatmap(vgm_t[[1]],feature.columns = c("VDJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3")
length(unique(vgm_t[[1]]$VDJ_cdr3s_aa))

clone_overlap <- VDJ_overlap_heatmap(vgm_t[[1]],feature.columns = c("VJ_cdr3s_aa","VDJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3")
length(unique(vgm_t[[1]]$pasted))

#make a table out of this for barplots (The numbers are from the length(unique())) prints above) 
to_plot <- data.frame(var = c("VDJ vgene","VJ vgene", "VDJ CDR3", "VJ CDR3", "VJ&VDJ CDR3"), ov = c(29,73,19,45,4), total = c(96,314,1546,1629,2094))
to_plot$fraction <- paste0(round((to_plot$ov / to_plot$total * 100),2),"%")

to_plot$var <- ordered(as.factor(to_plot$var),levels = c("VDJ vgene","VJ vgene", "VDJ CDR3","VJ CDR3", "VJ&VDJ CDR3"))

plot_out <- ggplot(data = to_plot, aes(x = var, y = ov, fill = ov)) + geom_bar(alpha = 1, stat = "identity") + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 30),axis.text.y = element_text(size = 27), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2),  axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25,"cm"), text = element_text(size=28), legend.position = "none", plot.margin = margin(l = 0.5, unit = "cm")) + labs(title = "Repertoire overlap", x = "", y = "shared elements") + geom_text(aes(label = fraction , vjust = -0.5), size = 9)+ scale_y_continuous(expand = c(0,0), limits = c(0,83))  + scale_fill_gradient(low = "grey80", high = stepped2(15)[13], oob = squish_infinite)

plot_out

#ggsave(plot_out, filename = "Tcs_rep_overlap_new.svg", dpi = 400, width = 8, height = 10)
```

To get the table of overlapping clones

```{r, fig.show='hold'}
print(head(clone_overlap[[3]]))
```

To visualize clone similarity across repertoires, we can also generate a similarity network. This can either filter out those unconnected clones (recommended for visualization purposes) or draw connections and show all nodes. Each node is a clone and the edge indicates that the distance between the two clones (based on CDR3) is less than the number supplied in the VDJ_network function

```{r, fig.show = 'hold'}
#get unique clones to avoid cluttering 
vgm_t[[1]]$sample_clone <- paste0(vgm_t[[1]]$sample_id, vgm_t[[1]]$clonotype_id_10x)
vgm_t_unique_clones <- subset(vgm_t[[1]][which(!duplicated(vgm_t[[1]]$sample_clone)),], clonotype_frequency > 5) #filtering: only one cell of each clone with a freq > 10
vgm_t_unique_clones$group_id <- as.character(vgm_t_unique_clones$group_id)
vgm_t_unique_clones$group_id[vgm_t_unique_clones$sample_id == "rMOG_1"] <- samples_color[1] #renewing the group id with the respective sample colors
vgm_t_unique_clones$group_id[vgm_t_unique_clones$sample_id == "rMOG_2"] <- samples_color[2]
vgm_t_unique_clones$group_id[vgm_t_unique_clones$sample_id == "MOG35-55"] <- samples_color[3]

#make a loop to plot for each sample a range of distance cutoff values
pdf(file = "Tcs_VDJ_network_filtered.pdf")
for(j in 3:12){ #here we generate a plot with different edit distance cutoffs to explore how the network behaves 
  netwo <- VDJ_network(VDJ = vgm_t_unique_clones,distance.cutoff = j,per.sample = F,platypus.version = "v3")
  igraph::plot.igraph(netwo[[4]],vertex.label=NA,vertex.size=5+(.03*as.numeric(netwo[[2]]$clonotype_frequency)),vertex.color=netwo[[2]]$group_id)
}
dev.off()

#for the purpose of this document, we view the plot at j = 9
netwo <- VDJ_network(VDJ = vgm_t_unique_clones,distance.cutoff = 9,per.sample = F,platypus.version = "v3")
igraph::plot.igraph(netwo[[4]],vertex.label=NA,vertex.size=5+(.03*as.numeric(netwo[[2]]$clonotype_frequency)),vertex.color=netwo[[2]]$group_id)
```

## Integrating VDJ and GEX sequencing information for T cells

The first step is to check how well these two datasets overlap. 

Nr. of cells in VDJ for which GEX info is available

```{r, fig.show='hold'}
GEX_av <- vgm_t[[1]] %>% group_by(sample_id) %>% summarize(av = sum(GEX_available), not_av = n()-sum(GEX_available)) #Can use the sum of this column as in R, TRUE is counted as numeric 1 and FALSE as numeric 0
GEX_av <- pivot_longer(GEX_av, cols = c(2,3))

plot_out <- list()
for(i in 1:3){

GEX_av_plot <- subset(GEX_av, sample_id == unique(GEX_av$sample_id)[i])
  
plot_out[[i]] <- ggplot(GEX_av_plot, aes(x = "", y=value, fill= name))+ geom_bar(stat = "identity", width = 1, show.legend = F)+ scale_fill_manual(values = c("darkorchid3","grey60"))+ coord_polar("y", direction = -1)+ theme(panel.background = element_blank(),axis.text.x = element_blank(),axis.text.y = element_blank(), axis.line = element_blank(), axis.ticks = element_blank(), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right",legend.direction = "vertical", plot.title = element_text(size = 20,hjust = 0.5)) + labs(x = "", y = "", title = paste0(unique(GEX_av$sample_id)[i])) + geom_text(x = 1, hjust = 1.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "not_av"], color = "white", fontface = "bold", size = 8) + geom_text(x = 1, hjust = -0.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "av"], color = "white", fontface = "bold", size = 8) + geom_point(inherit.aes = F, aes(x = 0, y = 1), color = "white", size = 25) 

#ggsave(plot_out[[i]], filename = paste0("GEX_available_pie_",plot_out[[i]]$labels$title ,".png"), dpi = 600, width = 5, height = 5)

}

print("Purple = Cells in VDJ for which GEX information is available")
print(plot_out)
print(GEX_av)
```

Nr. of cells in GEX for which VDJ info is available

```{r, fig.show='hold'}
VDJ_av <- vgm_t[[2]]@meta.data %>% group_by(sample_id) %>% summarize(av = sum(VDJ_available), not_av = n()-sum(VDJ_available)) 
VDJ_av <- pivot_longer(VDJ_av, cols = c(2,3))

plot_out <- list()
for(i in 1:3){

VDJ_av_plot <- subset(VDJ_av, sample_id == unique(VDJ_av$sample_id)[i])
  
plot_out[[i]] <- ggplot(VDJ_av_plot, aes(x = "", y=value, fill= name))+ geom_bar(stat = "identity", width = 1, show.legend = F)+ scale_fill_manual(values = c("darkorchid3","grey60"))+ coord_polar("y", direction = -1)+ theme(panel.background = element_blank(),axis.text.x = element_blank(),axis.text.y = element_blank(), axis.line = element_blank(), axis.ticks = element_blank(), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right",legend.direction = "vertical", plot.title = element_text(size = 20,hjust = 0.5)) + labs(x = "", y = "", title = paste0(unique(GEX_av$sample_id)[i])) + geom_text(x = 1, hjust = 1.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "not_av"], color = "white", fontface = "bold", size = 8) + geom_text(x = 1, hjust = -0.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "av"], color = "white", fontface = "bold", size = 8) + geom_point(inherit.aes = F, aes(x = 0, y = 1), color = "white", size = 25) 

#ggsave(plot_out[[i]], filename = paste0("VDJ_available_pie_",plot_out[[i]]$labels$title ,".png"), dpi = 600, width = 5, height = 5)

}

print("Purple = Cells in GEX for which VDJ information is available")
print(plot_out)
print(VDJ_av)
```

As a second step we now overlay VDJ information onto the GEX umap

```{r, fig.show='hold'}
#Add a discrete scale for expansion (To plot a continuos scale, use Seurat::FeaturePlot(..., features = "clonotype_frequency"))
vgm_t[[2]]$expanded <- vgm_t[[2]]$clonotype_frequency
vgm_t[[2]]$expanded[vgm_t[[2]]$clonotype_frequency > 1] <- "Expanded"
vgm_t[[2]]$expanded[vgm_t[[2]]$clonotype_frequency == 1] <- "1 cell"
vgm_t[[2]]$expanded[vgm_t[[2]]$VDJ_available == F] <- "VDJ not available"

plot_out <- Seurat::DimPlot(vgm_t[[2]],reduction = "umap", group.by = "expanded", cols = c("grey30","darkorchid3","grey80"), shuffle = T)+ theme(axis.text.y = element_blank(),axis.text.x = element_blank(),axis.line=element_blank(),axis.ticks = element_blank(), legend.position = "right") 
plot_ <- plot_out + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Clonal level", x = "UMAP1", y = "UMAP2") +guides(color=guide_legend(nrow=3,byrow=TRUE, override.aes = list(size=4)))

plot_ 

#ggsave(plot = plot_ , filename = "Tcs_GEXVDJ_expansion_discrete.svg", dpi = 700, width = 6.8, height = 5) 
```

Proportion barplots again help us visualize differential distribution among clusters

```{r, fig.show='hold'}
#adding a combination column
vgm_t[[2]]$sample_expanded <- paste0(vgm_t[[2]]$expanded, " ", vgm_t[[2]]$sample_id)
vgm_t[[2]]$sample_expanded[str_detect(vgm_t[[2]]$sample_expanded, "VDJ not available")] <- "VDJ not available" #Here we replace sample specific VDJ not available with a general one (e.g. rMOG_1 VDJ not available -> VDJ not available)
vgm_t[[2]]$sample_expanded <- ordered(as.factor(vgm_t[[2]]$sample_expanded), levels = c("Expanded rMOG_1", "1 cell rMOG_1","Expanded rMOG_2", "1 cell rMOG_2", "Expanded MOG35-55", "1 cell MOG35-55", "VDJ not available"))

plot_out <- GEX_proportions_barplot(GEX = vgm_t[[2]], source.group = "seurat_clusters", target.group = "sample_expanded", stacked.plot = T)

plot_out <-  plot_out + scale_fill_manual(values = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)])  + theme(panel.background = element_blank(),axis.text = element_text(size = 18), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=26), legend.key = element_rect(colour = "white"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), plot.margin = margin(l = 1, t = 1, unit = "cm"), legend.title = element_text(size = 18, face = "bold")) + labs(y = "% of cells of T cell group", x = "", fill = "Cluster")
  
plot_out

#ggsave(plot = plot_out, filename = paste0("Tcs_prop_sampleexpanded.svg"), dpi = 700, width = 10, height = 9) 
```

We are now ready to examine individual clones and their transcriptional phenotypes
The VDJ_clonal_expansion function gives great flexibility.

1. Expansion barplots by cluster (plotting by group ID given the large overlap between rMOG samples)
```{r, fig.show='hold'}
clonal_out <- VDJ_clonal_expansion(VDJ = vgm_t[[1]],celltype = "Tcells",clones = "50", platypus.version = "v3", group.by = "group_id", color.by = "seurat_clusters")

for(i in 1:length(clonal_out[[1]])){ #as this function will return a nested list of source dataframes and plots we loop over the output to refine the given plots
  
clonal_out[[1]][[i]] <- clonal_out[[1]][[i]] + theme(panel.background = element_blank(),axis.text.x = element_text(size = 25),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "bottom", plot.subtitle = element_text(size = 15), legend.title = element_text(size = 26, face = "bold")) + labs(title = unique(vgm_t[[1]]$group_id)[i]) + scale_fill_manual(values = c(pals::kelly(n = 17)[c(2,3,17,7,8,9,10,13,6,16)], "grey40","grey40")) + labs(fill = "Cluster", y = "n cells")+ ggplot2::scale_y_continuous(expand = c(0,0))

#ggsave(plot = clonal_out[[1]][[i]], filename = paste0("Tcs_cluster_per_clone_s_", unique(vgm_t[[1]]$group_id)[i],".svg"), dpi = 400, width = 11, height = 7)
}

clonal_out[[1]]
```

2. Expansion barplots by T cell subtype

```{r, fig.show='hold'}
#adding a column for T cell group
vgm_t[[2]]$Tc_group <- "Other"
vgm_t[[2]]$Tc_group[which(GetAssayData(vgm_t[[2]], assay = "RNA", slot = "counts")[which(rownames(vgm_t[[2]]) == "CD4"),] > 0)] <- "CD4 Th"
vgm_t[[2]]$Tc_group[which(GetAssayData(vgm_t[[2]], assay = "RNA", slot = "counts")[which(rownames(vgm_t[[2]]) == "CD8A"),] > 0)] <- "CD8 CTL"
vgm_t[[2]]$Tc_group[which(GetAssayData(vgm_t[[2]], assay = "RNA", slot = "counts")[which(rownames(vgm_t[[2]]) == "FOXP3"),] > 0)] <- "FOXP3 Treg"

vgm_t[[2]]$Tc_group <- ordered(as.factor(vgm_t[[2]]$Tc_group),levels = c("CD4 Th", "CD8 CTL", "FOXP3 Treg", "Other"))

#We transfer this extra T cell group column also to the VDJ matrix vgm_t[[1]] by barcode matching
vgm_t[[1]]$Tc_group <- "Other"
for(i in 1:3){
  vgm_t[[1]]$Tc_group[which(vgm_t[[1]]$barcode %in% colnames(vgm_t[[2]])[which(vgm_t[[2]]$Tc_group == unique(vgm_t[[2]]$Tc_group)[i])])] <- as.character(unique(vgm_t[[2]]$Tc_group)[i])
}
#just checking that it worked
unique(vgm_t[[1]]$Tc_group)
table(vgm_t[[1]]$Tc_group)
table(vgm_t[[2]]$Tc_group[which(vgm_t[[2]]$VDJ_available)])

clonal_out <- VDJ_clonal_expansion(VDJ = vgm_t[[1]],celltype = "Tcells",clones = "50", platypus.version = "v3", group.by = "group_id", color.by = "Tc_group")

for(i in 1:length(clonal_out[[1]])){
  
clonal_out[[1]][[i]] <- clonal_out[[1]][[i]] + theme(panel.background = element_blank(),axis.text.x = element_text(size = 25),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "bottom", plot.subtitle = element_text(size = 15), legend.title = element_text(size = 26, face = "bold")) + labs(title = unique(vgm_t[[1]]$group_id)[i]) + scale_fill_manual(values = c(pals::stepped(n = 16)[c(6,1,9)],"grey80"), labels = c("CD4+ Th", "CD8+ CTL", "FOXP3+ Treg", "Other")) + labs(fill = "Subtype", y = "n cells")+ ggplot2::scale_y_continuous(expand = c(0,0))+guides(fill=guide_legend(nrow=3,byrow=TRUE))

#ggsave(plot = clonal_out[[1]][[i]], filename = paste0("Tcs_celltype_per_clone_s_", unique(vgm_t[[1]]$group_id)[i],".svg"), dpi = 400, width = 11, height = 7)
}

clonal_out[[1]]
```

Proceeding to DEG analysis by expansion at T cell subtype level

```{r, fig.show='hold'}
#we join the expanded and T cell group to get a more detailed column. Then we use this column as input for the DEG genes analysis and find what sets these groups apart
vgm_t[[2]]$Tc_group_expanded <- paste0(vgm_t[[2]]$expanded, " ", vgm_t[[2]]$Tc_group)


vgm_t[[2]]$Tc_group_expanded <- ordered(as.factor(vgm_t[[2]]$Tc_group_expanded), levels = c("Expanded CD4 Th", "1 cell CD4 Th","Expanded CD8 CTL", "1 cell CD8 CTL","Expanded FOXP3 Treg", "1 cell FOXP3 Treg", "Expanded Other", "1 cell Other","VDJ not available CD4 Th", "VDJ not available CD8 CTL", "VDJ not available FOXP3 Treg", "VDJ not available Other"))

#Showing numbers
table(vgm_t[[2]]$Tc_group_expanded)
```

First helper CD4+ cells (Th)

```{r, fig.show='hold'}
exp_DEGs <- GEX_DEgenes(vgm_t[[2]], min.pct = 0.25, grouping.column = "Tc_group_expanded",group1 = "Expanded CD4 Th", group2 = "1 cell CD4 Th", return.plot = "volcano", logFC = F) 

#plot volcano
plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 1)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "Expanded vs. 1 cell CD4+") + geom_text_repel(data = subset(exp_DEGs[[1]], -log10(p_val_adj) > 8 | SYMBOL %in% c("TCF7", "GZMB", "CD28", "CD69", "HIF1A", "TOX")), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 40)+scale_color_gradient2(low = "grey10", mid = "grey60", high = "darkorchid3", midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_volc_cd4_expanded.svg", dpi = 400, width = 9, height = 7)
```

Next: CD8+ CTLs

```{r, fig.show='hold'}
exp_DEGs <- GEX_DEgenes(vgm_t[[2]], min.pct = 0.25, grouping.column = "Tc_group_expanded",group1 = "Expanded CD8 CTL", group2 = "1 cell CD8 CTL", return.plot = "volcano", logFC = F)

#plot volcano
plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "Expanded vs. 1 cell CD8+") + geom_text_repel(data = subset(exp_DEGs[[1]], -log10(p_val_adj) > 1.8 | SYMBOL %in% c("TCF7", "GZMB", "CD28", "CD69", "TOX", "JUN")), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 40)+scale_color_gradient2(low = "grey10", mid = "grey60", high = "darkorchid3", midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_volc_cd8_expanded.svg", dpi = 400, width = 9, height = 7)
```

And finally Tregs

```{r, fig.show='hold'}
exp_DEGs <- GEX_DEgenes(vgm_t[[2]], min.pct = 0.25, grouping.column = "Tc_group_expanded",group1 = "Expanded FOXP3 Treg", group2 = "1 cell FOXP3 Treg", return.plot = "volcano", logFC = F)

#plot volcano
plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 1)  + theme(panel.background = element_blank(),axis.text = element_text(size = 23), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=23), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "Expanded vs. 1 cell FOXP3+") + geom_text_repel(data = subset(exp_DEGs[[1]], -log10(p_val_adj) > 0.4 | abs(avg_logFC) > 0.95), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 40)+scale_color_gradient2(low = "grey10", mid = "grey60", high = "darkorchid3", midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Tc_volc_foxp3_expanded.svg", dpi = 400, width = 9, height = 7)
```

## T cells VDJ GEX overview by ProjecTILs 

A recent publication (Andreatta, M. et al. Interpretation of T cell states from single-cell transcriptomics data using reference atlases. Nat. Commun. 12, 2965 (2021)) put forward a reference atlas of murine T cells, onto which local single cell data can easily be projected. We use this to verify our observation about phenotypes of expanded T cells
Much of this is copied from the papers vignette, which we highly recommend. 

```{r, fig.show='hold', include = FALSE}
rm(vgm_EAEaged)
rm(vgm_EAEaged_allGEX)
gc()
```


```{r, fig.show='hold'}
library(remotes)
library(ProjecTILs) #For installation instructions please refer to github carmonalab/ProjecTILs

#ref <- load.reference.map() #If you do not have the reference dataset downloaded, it will download this for you. Loading is done via the same command. 
#For me loading did not work as expected, so we downloaded the file manually via my browser (https://doi.org/10.6084/m9.figshare.12489518) and read it in
ref <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/ref_TILAtlas_mouse_v1.rds")

#Generating a dimplot of the reference 
refCols <- c("#edbe2a", "#A58AFF", "#53B400", "#F8766D", "#00B6EB", "#d1cfcc", "#FF0000", "#87f6a5", "#e812dd")
DimPlot(ref,label = T, cols = refCols)

#And feature plots of the reference
markers <- c("Cd4","Cd8a","Ccr7","Tcf7","Pdcd1","Havcr2","Tox","Izumo1r","Cxcr6","Xcl1","Gzmb","Gzmk","Ifng","Foxp3")
VlnPlot(ref,features=markers,stack = T,flip = T,assay = "RNA")
```

Overlaying the query data (vgm_t[[2]]) to this reference

```{r, fig.show='hold'}
#make query projection
query.projected <- make.projection(vgm_t[[2]], ref=ref)

#Plot overlay on umap as density
plot.projection(ref, query.projected)
#ggsave(plot = last_plot(), filename = "ProjecTILs_overlay.png", dpi = 700, width = 8, height = 6)

#Get a barplot of cell states
query.projected <- cellstate.predict(ref=ref, query=query.projected)
plot.statepred.composition(ref, query.projected,metric = "Percent")
#ggsave(plot = last_plot(), filename = "ProjecTILs_celltypes.png", dpi = 700, width = 6, height = 5)
```

We can then use the expansion information to check for reference T cell phenotypes that match our expanded phenotype

```{r, fig.show='hold'}
#split by that
not_expanded <- subset(query.projected, expanded == "1 cell")
expanded <- subset(query.projected, expanded == "Expanded")

plot.projection(ref, expanded)

plot.projection(ref, not_expanded)

#summarize this and plot the same as an easy to view stacked barplot
melting <- query.projected@meta.data %>% group_by(expanded, functional.cluster) %>% summarise(n = n())
for_norm <- query.projected@meta.data %>% group_by(functional.cluster) %>% summarise(n_all = n())
melting <- merge(melting,for_norm, by ="functional.cluster", all.x = T)
melting$norm <- melting$n / melting$n_all * 100 #normalize frequencies across reference clusters

output.plot <- ggplot(melting, aes(fill = expanded, y=norm, x= functional.cluster)) + geom_bar(stat="identity", width=0.6, color="black", position = "stack")  + ylab(paste0("% of cells of clonal group")) + xlab("ProjectTIL phenotype assignment") + scale_fill_manual(values = c("grey30","darkorchid3","grey70"))  + theme(panel.background = element_blank(),axis.text = element_text(size = 18), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=26), legend.key = element_rect(colour = "white"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), plot.margin = margin(l = 1, t = 1, unit = "cm"), legend.title = element_text(size = 18, face = "bold")) + labs(fill = "Clonal level")

output.plot

#ggsave(plot = output.plot, filename = paste0("ProjecTILs_stacked_expansion.png"), dpi = 700, width = 9, height = 9) 
```

```{r, include = FALSE}
rm(ref)
rm(Hs2Mm.convert.table)
rm(not_expanded)
rm(list = ls(pattern = "VDJ_"))
rm(query.projected)
rm(expanded)
gc() #clear some ram
```

## T cells VDJ GEX integration of human data by Beltrán et al. 

A second recent publication (Beltrán, E. et al. Early adaptive immune activation detected in monozygotic twins with prodromal multiple sclerosis. J. Clin. Invest. 129, 4758–4768 (2019)) about lymphocytes in the CNS of discordant twins (twins were one developed Multiple sclerosis, while the other has not) highlighted a resident memory-like phenotype of expanded T cells in MS as well as other inflammatory neurological conditions (SCNwe = subclinical neuroinflammation, Enc = autoimmune encephalitis, HD /NIC = Healthy donors, Non-inflammatory controls)

Here we load in the raw data of that publication to investigate if similar cellular patterns can be observed in murine EAE

```{r, fig.show='hold'}
#data downloaded and loaded loaded in manually from GEO GSE127969
#tpm <- read.delim("GSE127969_counts_TPM_ALL.csv.gz")
inf <- read.delim("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/GSE127969_sc_cell_info.txt.gz")
#bar <-  read.delim("GSE127969_cellID_barcode_SRR.txt.gz")

#make Seurat object
#rownames(tpm) <- tpm$Symbol
#tpm <- tpm[,-c(1)]

#tpm_s <- CreateSeuratObject(tpm, names.delim = "/", names.field = 1)
#tpm <- NULL

#using a Platypus function to process raw reads
#tpm_s <- automate_GEX(GEX.list = list(tpm_s), GEX.outs.directory.list = "none")

#saveRDS(tpm_s, "Beltran_GEX.rds")

#Loading in a processed object for faster knitting
tpm_s <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/Beltran_GEX.rds")

#figure out which cells are which
inf_names <- str_extract(inf$Cell, "_[A-Z]\\d\\d")
tpm_names <- str_extract(colnames(tpm_s), "_[A-Z]\\d+")

#apparently there are a few more cells in info than in the object. Lets remove those
met <- tpm_s[[1]]@meta.data
met$Cell <- rownames(met)
new_met <- merge(met, inf, by = "Cell", all.x = T)
rownames(new_met) <- new_met$Cell
tpm_s <- AddMetaData(tpm_s[[1]], new_met) #make tpm_s a Seurat object instead of a list

#add expansion information based on clone column
tpm_s$Expanded <- "1 cell"
tpm_s$Expanded[which(tpm_s$Clones > 1 & tpm_s$Clones != "ND")] <- "Expanded"
tpm_s$exp_past <- paste0(tpm_s$Expanded, " ", tpm_s$index.sort)
tpm_s$exp_past_diag <- paste0(tpm_s$Expanded, " ", tpm_s$Case)

#SUBSET CD8
tpm_cd8 <- subset(tpm_s, index.sort == "CD8Tcell")
table(tpm_cd8$exp_past_diag )

#order by condition and clonal level
tpm_cd8$exp_past_diag <- ordered(as.factor(tpm_cd8$exp_past_diag), levels = c("1 cell HD", "Expanded HD","1 cell SCNI", "Expanded SCNI", "1 cell MS", "Expanded MS", "1 cell Enc", "Expanded Enc"))

#plot dotmap
linesv <- seq.int(0,10,2) + 0.5
plot_out <- GEX_dottile_plot(tpm_cd8, genes = c("CD4", "CD8A","PTPRC","CCR7","SELL", "CD69","CXCR6", "CCR5","TCF7","S1PR1", "IL7R","TOX", "SOCS3","GZMB","TGFB1"), group.by = "exp_past_diag")+ labs(title = "Human CD8+ T cells") +geom_vline(xintercept = linesv, color = "black", size= 0.9) + scale_size_binned(range = c(1,13))

#for showing in the vignette
plot_out + theme(legend.position = "none")

#ggsave(plot_out, filename = "TcsBeltran_dottile_CD8_rel.svg", width = 11.5, height = 11.5, dpi = 300)

#SUBSET CD4
tpm_cd4 <- subset(tpm_s, index.sort == "CD4Tcell")
table(tpm_cd4$exp_past_diag )

#order by condition and clonal level
tpm_cd4$exp_past_diag <- ordered(as.factor(tpm_cd4$exp_past_diag), levels = c( "1 cell HD", "Expanded HD","1 cell SCNI", "Expanded SCNI", "1 cell MS", "Expanded MS", "1 cell Enc", "Expanded Enc"))

#plot dotmap
linesv <- seq.int(0,10,2) + 0.5
plot_out <- GEX_dottile_plot(tpm_cd4, genes = c("CD4", "CD8A","PTPRC","CCR7","SELL", "CD69","CXCR6", "CCR5","TCF7","S1PR1", "IL7R","TOX", "SOCS3","GZMB","TGFB1"), group.by = "exp_past_diag") + labs(title = "Human CD4+ T cells") + theme(axis.text.x= element_text(vjust = 1, hjust = 1))+geom_vline(xintercept = linesv, color = "black", size= 0.9)+ scale_size_binned(range = c(1,13))

#for showing in the vignette
plot_out + theme(legend.position = "none")

#ggsave(plot_out, filename = "TcsBeltran_dottile_CD4_rel.svg", width = 10, height = 10, dpi = 300)
```

```{r, fig.show='hold'}
rm(bar)
rm(inf)
rm(met) 
rm(list=ls(pattern="tpm"))
gc()
```

We follow this up with identical dottiles containing EAE and aged CNS data

```{r, fig.show='hold', include=FALSE}
vgm_EAEaged <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_EAEaged.rds")
```


```{r, fig.show='hold'}
#add expansion measures
vgm_EAEaged[[2]]$expanded <- "VDJ not available"
vgm_EAEaged[[2]]$expanded[vgm_EAEaged[[2]]$clonotype_frequency > 1] <- "Expanded"
vgm_EAEaged[[2]]$expanded[vgm_EAEaged[[2]]$clonotype_frequency == 1] <- "1 cell"
vgm_EAEaged[[2]]$expanded[vgm_EAEaged[[2]]$VDJ_available == F] <- "VDJ not available"

#we join the expanded and T cell group to get a more detailed column. Then we use this column as input for the DEG genes analysis and find what sets these groups apart
#t cell group
vgm_EAEaged[[2]]$Tc_group <- "Other" 
vgm_EAEaged[[2]]$Tc_group[which(GetAssayData(vgm_EAEaged[[2]], assay = "RNA", slot = "counts")[which(rownames(vgm_EAEaged[[2]]) == "CD4"),] > 0)] <- "CD4"
vgm_EAEaged[[2]]$Tc_group[which(GetAssayData(vgm_EAEaged[[2]], assay = "RNA", slot = "counts")[which(rownames(vgm_EAEaged[[2]]) == "CD8A"),] > 0)] <- "CD8"
vgm_EAEaged[[2]]$Tc_group[which(GetAssayData(vgm_EAEaged[[2]], assay = "RNA", slot = "counts")[which(rownames(vgm_EAEaged[[2]]) == "FOXP3"),] > 0)] <- "FOXP3"

vgm_EAEaged[[2]]$group_expanded <- paste0(vgm_EAEaged[[2]]$expanded, " ", vgm_EAEaged[[2]]$group_id)
table(vgm_EAEaged[[2]]$group_expanded)

vgm_EAEaged_allGEX <- subset(vgm_EAEaged[[2]], expanded != "VDJ not available") #removing cells without VDJ information from GEX
vgm_EAEaged_allGEX$group_expanded <- ordered(as.factor(vgm_EAEaged_allGEX$group_expanded), levels = c("1 cell EAE", "Expanded EAE", "1 cell aged CNS", "Expanded aged CNS"))

#dottile CD8
plot_out <- GEX_dottile_plot(subset(vgm_EAEaged_allGEX, Tc_group == "CD8"), genes = c("CD4", "CD8A","PTPRC","CCR7","SELL", "CD69","CXCR6", "CCR5","TCF7","S1PR1", "IL7R","TOX", "SOCS3","GZMB","TGFB1"), group.by = "group_expanded")+ labs(title = "Murine CD8+ T cells")+ scale_size_binned(range = c(1,13))+geom_vline(xintercept = c(1.5,2.5,3.5), color = "black", size= 0.9)

plot_out + theme(legend.position = "none")

#ggsave(plot_out, filename = "TcsEAEaged_dottile_CD8_rel.svg", width = 10, height = 12, dpi = 300)


#dottile CD4
plot_out <- GEX_dottile_plot(subset(vgm_EAEaged_allGEX, Tc_group == "CD4"), genes = c("CD4", "CD8A","PTPRC","CCR7","SELL", "CD69","CXCR6", "CCR5","TCF7","S1PR1", "IL7R","TOX", "SOCS3","GZMB","TGFB1"), group.by = "group_expanded")+ labs(title = "Murine CD4+ T cells") + scale_size_binned(range = c(1,13))+geom_vline(xintercept = c(1.5,2.5,3.5), color = "black", size= 0.9)

plot_out + theme(legend.position = "none")

#ggsave(plot_out, filename = "TcsEAEaged_dottile_CD4_rel.svg", width = 10, height = 12, dpi = 300)
```

As a final view of T cells in EAE and aged CNS, we can replot the umap from before, but overlay clonal expansion information

```{r, fig.show='hold'}
#Add group expansion identifier column
vgm_EAEaged[[2]]$group_expanded <- paste0(vgm_EAEaged[[2]]$group_id, " ", vgm_EAEaged[[2]]$expanded)
table(vgm_EAEaged[[2]]$group_expanded)

vgm_EAEaged[[2]]$group_expanded <- ordered(as.factor(vgm_EAEaged[[2]]$group_expanded), levels = c("EAE 1 cell", "EAE Expanded", "aged CNS 1 cell", "aged CNS Expanded", "EAE VDJ not available", "aged CNS VDJ not available"))

plot_out <- Seurat::DimPlot(vgm_EAEaged[[2]],reduction = "umap", group.by = "group_expanded", cols = c("grey30","darkorchid3", samples_color_aged[[5]], "gold2", "grey80" ,"grey80"), shuffle = T, pt.size = 1.1)+ theme(axis.text.y = element_blank(),axis.text.x = element_blank(),axis.line=element_blank(),axis.ticks = element_blank(), legend.position = "bottom") 
plot_out <- plot_out + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Clonal level", x = "UMAP1", y = "UMAP2") +guides(color=guide_legend(nrow=3,byrow=TRUE, override.aes = list(size=4)))

plot_out 

#ggsave(plot = plot_ , filename = "Tcs_EAEaged_expansion_discrete.svg", dpi = 700, width = 10, height = 5) 
```

```{r, fig.show='hold', include=FALSE}
rm(vgm_EAEaged)
rm(vgm_EAEaged_allGEX) 
gc()


source_code_dir <- "C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/PlatypusDB_dev_VK/R" 
file_path_vec <- list.files(source_code_dir, full.names = T)
for(f_path in file_path_vec){
  print(f_path)
  tryCatch({source(f_path)}, error = function(e){e})
}

#load in for the next chunk
comb <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/Combined VDJ_comb Literature MOG TCRs.rds")

```

## T cells MOG-specific TCRs by public dataset integration

A key goal was to validate the presence of MOG-specific T cells within our dataset. 
For this a literature search resulted in several T cell repertoire datasets in EAE. 
For dataset reference see publication

As part of this vignette we do not show the whole download and formatting process of these, but rather load in a formatted table of CDR3s

```{r, fig.show='hold', warning=FALSE}
#comb <- readRDS("~/Combined VDJ_comb Literature MOG TCRs.rds")

unique(comb$group_id)

#minor corrections in naming scheme
comb$group_id <- gsub("MOG-specific ", "", comb$group_id)
comb$group_id <- gsub("Aged", "aged CNS", comb$group_id)

#addin a barcode column that the overlap function needs to generate an adjunct frequency table with tidy information
comb$barcode <- c(1:nrow(comb))

cdra_overlap <- VDJ_overlap_heatmap(VDJ = comb,
                    feature.columns = "VJ_cdr3s_aa",
                    grouping.column = "group_id",
                    pvalues.label.size = 6,
                    axis.label.size = 16,
                    add.barcode.table = T,
                    platypus.version = "v3",
                    plot.type = "pheatmap")


cdra_overlap[[1]]

#ggsave(last_plot(), filename = "MOGspec_Overlap_cdra.svg", dpi = 400, width = 6.5, height = 6.5)


cdrb_overlap <- VDJ_overlap_heatmap(VDJ = comb,
                    feature.columns = "VDJ_cdr3s_aa",
                    grouping.column = "group_id",
                    pvalues.label.size = 6,
                    axis.label.size = 16,
                    add.barcode.table = T,
                    platypus.version = "v3",
                    plot.type = "pheatmap")

cdrb_overlap[[1]]

#ggsave(cdrb_overlap[[1]], filename = "MOGspec_Overlap_cdrb.svg", dpi = 400, width = 6.5, height = 6.5)
```

We found two clones in the EAE MOG35-55 dataset that were reported to be MOG35-55-specific by Saligrama et al. Here we briefly explore the phenotypes of these cells

```{r, fig.show='hold'}
#add specificity info to VDJ...
vgm_t[[1]]$MOG_spec <- F
vgm_t[[1]]$MOG_spec[which(vgm_t[[1]]$VDJ_cdr3s_aa %in% c("CASGDAGGGQNTLYF", "CASSLETANTEVFF") | vgm_t[[1]]$VJ_cdr3s_aa %in% c("CAASEGQGGRALIF", "CAVSAGTNAYKVIF"))] <- T

#...and GEX table
vgm_t[[2]]$MOG_spec <- vgm_t[[2]]$clonotype_id_10x
vgm_t[[2]]$MOG_spec[which(!vgm_t[[2]]$VDJ_cdr3s_aa %in% c("CASGDAGGGQNTLYF", "CASSLETANTEVFF") | !vgm_t[[2]]$VJ_cdr3s_aa %in% c("CAASEGQGGRALIF", "CAVSAGTNAYKVIF"))] <- "FALSE"

#Standard dotmap first 
genes_to_plot <- c("CD4","CD8A","CD28","ICOS","CD40LG","PDCD1","LAG3","S1PR1","PTPRC","SELL","TBX21","GATA3","SPI1","IRF4","BCL6","STAT4","STAT6","FOXP3", "MKI67","TCF7","TOX", "GZMB","IFNG","TGFB1","IL10","IL17A","CSF2","IL2RA","IL4RA","IL12RB1","CXCR3","CCR5","CCR8","ITGA4","ITGB2")

linesh <- c(0:length(genes_to_plot)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_t[[2]], genes = genes_to_plot, group.by = "MOG_spec", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "MOG specific") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 45, hjust = 1,vjust = 1)) +scale_size_binned(range = c(0.01,12)) + geom_hline(yintercept = linesh, col = "black", size = 0.9)+ scale_color_viridis_c(option = "B", end = 0.85)

#for showing in the vignette
plot_out + theme(legend.position = "none")

#ggsave(plot = plot_out, filename = "MOGspec_cluster_custom_genes_dottile.svg", dpi = 400, width = 8, height = 14.5)

#project onto umap
vgm_t[[2]]$MOG_spec_umap <- FALSE
vgm_t[[2]]$MOG_spec_umap[vgm_t[[2]]$MOG_spec != "FALSE"] <- T

plot_out <- VDJ_GEX_overlay_clones(GEX = vgm_t[[2]],reduction = "umap", by.sample = F, clones.to.plot = "MOG_spec_umap",ncol.facet = 1,pt.size = 0.6,split.plot.and.legend = T, platypus.version = "v3", by.other.group = F)

#Package needed below: 
library(gridExtra)
library(cowplot)

grid.arrange(plot_out[[2]]) #as a grid object, we need to use the plot function to print the legend
plot_out[[1]]

#ggsave(plot = plot_out[[2]], filename = "MOGspec_overlay_leg.svg", dpi = 700, width = 10, height = 10)
#ggsave(plot = plot_out[[1]], filename = "MOGspec_overlay_plot.svg", dpi = 700, width = 4.5, height = 4.5) 
```

```{r, include=F}
rm(list = ls(pattern = "cdra_"))
comb <- NULL
new_met <- NULL
plot_ <- NULL
plot_out <- NULL
vgm_t <- NULL
gc()
```

This concludes our journey through the T cells from this dataset. We will now progress on to B cells

## Exploring GEX of B cells by cluster

The walk through of B cells will in many ways be very similar to that of T cells. The VDJ and VDJ GEX integration contains more specific B cell-specific analysis.

Querying the transcriptional phenotypes of B cells in this dataset. 
We will start with some overview UMAPs and work towards cluster classification

```{r, fig.show='hold'}

#By sample 
Seurat::DimPlot(vgm_b[[2]],reduction = "umap", group.by = "sample_id", cols = samples_color, shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Sample", x = "UMAP1", y = "UMAP2") 

#ggsave(plot = last_plot(), filename = "Bcs_umap_sample.svg", dpi = 700, width = 6.5, height = 5)

#By group
Seurat::DimPlot(vgm_b[[2]],reduction = "umap", group.by = "group_id", cols = groups_color, shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Sample", x = "UMAP1", y = "UMAP2") 

#ggsave(plot = last_plot(), filename = "Bcs_umap_group.svg", dpi = 700, width = 6.5, height = 5)

#By cluster
library(pals) #the pals package offers some great varities of palettes, especially for applications were rColorBrewer palettes contain too few distinct colors

Seurat::DimPlot(vgm_b[[2]],reduction = "umap", group.by = "seurat_clusters", cols = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)], shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Cluster", x = "UMAP1", y = "UMAP2") 

#ggsave(plot = last_plot(), filename = "Bcs_umap_clusters.svg", dpi = 700, width = 6.5, height = 5)
```

We see clear overlap between samples. Almost all clusters are composed of cells of all samples and groups. 
To visualize proportions specifically, we use a stacked barplot

```{r, fig.show='hold'}
cl_members <- GEX_proportions_barplot(GEX = vgm_b[[2]],source.group = "sample_id", target.group = "seurat_clusters", stacked.plot = T)+ scale_fill_manual(values = samples_color, labels = short_replace_ids) + theme(panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "bottom", plot.title = element_blank(), legend.title = element_text(size = 30, face = "bold")) + labs(title = "", x = "", y = "% of cells of cluster", fill = "Sample") + scale_y_continuous(expand = c(0,0))

cl_members 

#ggsave(last_plot(), filename = "Bc_cluster_membership_sample_new.svg", dpi = 400, width = 10, height = 8)
```

Checking what the cluster represent. 
Step 1: Project common markers to get a basic idea

```{r, fig.show='hold'}
library(scales)

FeaturePlot(vgm_b[[2]], features = c("CD19","PTPRC", "CD74", "MS4A1"), reduction = "umap", label = F, repel = F, pt.size = 1.3) & theme(axis.text.y = element_blank(),axis.text.x = element_blank(),axis.line=element_line(size = 1),axis.ticks = element_blank(), legend.position = c(0.5,0.1), legend.direction = "horizontal", axis.title = element_text(), plot.title = element_text(size = 27, face = "bold")) & labs(x = "UMAP1", y = "UMAP2") & scale_color_gradient2(low = "grey80", mid = "grey80", high = "#DD513AFF", oob = squish_infinite) #We position the legend inside the plot using legend.position, to save some space. Further we use the "squish_infinite" function of the scales package to easily reppresent outlier values without compromising overall color scales.  

#ggsave(plot = last_plot(), filename = "Bcs_umap_Tcell_basic.svg", dpi = 400, width = 10, height = 10)
```

For a more in depth look at marker expression, we use the dottile function for some common B cell markers.

```{r, fig.show='hold'}
#Check what genes are available with this searchable dataframe view (at least in R studio)
#View(as.data.frame(rownames(vgm_b)))

genes_to_plot <- c("CD19", "CD74","SDC1", "EBF1","PTPRC","CD38","CD24A","CD1D1","CD5","MS4A1","CXCR5","SELL","CD40","CD83","H2-AB1","H2-EB1","CD27","POU2AF1","NT5E","FASL","PDCD1LG2","PRDM1","ITGAM","TGFB1","ZCWPW1")

linesh <- c(1:length(genes_to_plot)) + 0.5 #these lines are for easier viewing of the plot
plot_out <- GEX_dottile_plot(GEX = vgm_b[[2]], genes = genes_to_plot, group.by = "seurat_clusters", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 0, hjust = 0.5,vjust = 0.5)) +scale_size_binned(range = c(0.01,12)) + geom_hline(yintercept = linesh, col = "black", size = 0.9)+ scale_color_viridis_c(option = "B", end = 0.85) #we end the viridis palette at .85 specifically, to avoid bright yellow spots that can be hard to view when printed

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = last_plot(), filename = "Bc_dottile.svg", dpi = 400, width = 10, height = 15)
```

After getting some visual evidence the search leads to cluster specific genes by DEG analysis. 
Here we are using the GEX_cluster_genes to find genes that are deferentially expressed in one cluster in comparison to all other cluster

```{r, fig.show='hold'}
## Warning: running this function will take a while
gene_expression_cluster <- GEX_cluster_genes(vgm_b[[2]],min.pct = 0.25) 

#saveRDS(gene_expression_cluster, file = "Bcs_gene_expression_clusters.rds")

#Recommended to get an overview of the output format
names(gene_expression_cluster[[1]])
#View(gene_expression_cluster[[1]])
```

We can visualize these cluster defining genes both as a heatmap, a dotmap and as volcano plots

```{r, fig.show='hold'}
#heatmap
EAE_heatmap_clusters <- GEX_cluster_genes_heatmap(GEX = vgm_b[[2]], GEX_cluster_genes.output = gene_expression_cluster, n.genes.per.cluster = 5,max.cell = 30, metric = "top_logFC", group.colors = kelly(n = 17)[c(2,3,17,7:13,6,16)]) + scale_fill_gradientn(colors = c("white","grey80","darkorchid3"), oob = squish_infinite) + theme(axis.text.y = element_text(size = 13),axis.text.x = element_blank(), legend.position = "none") #here we picked top genes by fold change value. It is recommended to also check top genes by p value for consistency

EAE_heatmap_clusters

#ggsave(plot = last_plot(), filename = "Bc_cluster_heatmap_top5_logFC.svg", dpi = 400, width = 10, height = 10)
```

The dottile function can be run directly from the gene selection that made by the heatmap function

```{r, fig.show='hold'}
cldef <- EAE_heatmap_clusters$data #getting the source data table of the heatmap from above
length(unique(cldef$Feature)) #how many genes were selected

genes_to_plot <- as.character((unique(cldef$Feature)))
lines <- c(seq(0,58,5)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_b[[2]], genes = rev(genes_to_plot), group.by = "seurat_clusters", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 0, hjust = 0.5,vjust = 0.5))  +scale_size_binned(range = c(0.01,12)) + geom_hline(yintercept = lines, col = "black", size = 0.9) + guides(color = guide_legend(reverse=T))+ scale_color_viridis_c(option = "B", end = 0.85)

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Bc_cluster_dottile_top5_logFC.svg", dpi = 400, width = 10, height = 22)

```

And some more DEG analysis between clusters: 
We notice a difference between the two sides of the UMAP. Here we look at this specifically 

```{r, fig.show='hold', eval = FALSE}
vgm_b[[2]]$umap_side <- "none"
vgm_b[[2]]$umap_side[vgm_b[[2]]$seurat_clusters %in% c(1,3)] <- "left"
vgm_b[[2]]$umap_side[vgm_b[[2]]$seurat_clusters %in% c(0,2,4)] <- "right"

table(vgm_b[[2]]$umap_side)#Check cell nr.
#Get DEGs
exp_DEGs <- GEX_DEgenes(vgm_b[[2]], min.pct = 0.25, grouping.column = "umap_side", group1 = "right", group2 = "left", return.plot = "volcano", logFC = F, label.n.top.genes = 20, genes.to.label = c("CD24A", "BACH2", "EBF1", "CD83", "PTPRC", "CD38", "ARHGAP15", "MALAT1", "DOCK2", "FOXP1", "LNCPINT", "TGFB1"))

#plot volcano
plot_out <- exp_DEGs[[2]] + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "UMAP lobe right vs. left") +scale_color_gradient2(low = "grey30", mid = "grey30", high = "grey60", midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Bcs1_volc_umap_rightvleft.svg", dpi = 400, width = 9.5, height = 6.5)
```

## B cell GEX by group

A key goal of this study is the comparison between B cells in EAE initiated by two different autoantigens: rMOG and MOG35-55
Here we compare B cells from the two conditions. 

```{r, fig.show='hold'}

#first a membership barplot to check if B cells from either immunizations cluster differently
cl_members <- GEX_proportions_barplot(GEX = vgm_b[[2]],source.group = "group_id", target.group = "seurat_clusters", stacked.plot = T)+ scale_fill_manual(values = groups_color, labels = c("rMOG","MOG35-55")) + theme(panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "bottom", plot.title = element_blank(), legend.title = element_text(size = 30, face = "bold")) + labs(title = "", x = "", y = "% of cells of cluster", fill = "Sample") + scale_y_continuous(expand = c(0,0))

cl_members 

#ggsave(last_plot(), filename = "Bc_cluster_membership_group.svg", dpi = 400, width = 10, height = 8)

#Get DEGs 
exp_DEGs <- GEX_DEgenes(GEX = vgm_b[[2]], min.pct = 0.25, grouping.column = "group_id", group1 = "rMOG", group2 = "MOG35-55", return.plot = "none", logFC = F) #here we separate the function call from the ggplot for more flexibility. output[[1]] of the GEX_DEgenes is a dataframe with DEGs

#remove the max fold change gene. Most likely an artifact
exp_DEGs[[1]] <- exp_DEGs[[1]][rownames(exp_DEGs[[1]]) != "GM30211",]

plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 1)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "rMOG vs. MOG33-55") + geom_text_repel(data = subset(exp_DEGs[[1]], p_val_adj < 0.05), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 20)+scale_color_gradient2(low = samples_color[2], mid = "grey60", high = samples_color[1], midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Bc_volc_bygroup.svg", dpi = 400, width = 13, height = 11)

#DEGs after p correction
length(exp_DEGs[[1]]$p_val_adj[exp_DEGs[[1]]$p_val_adj < 0.05])

```

Running Gene ontology analysis
This outputs plots as PDFs, which are not displayed in this vignette

```{r, fig.show='hold'}
#ontology_EAE <- GEX_GOterm(GEX.cluster.genes.output = list(exp_DEGs[[1]]), topNgenes = 40, go.plots = T)
```

For completeness we also plot a dotmap of selected genes across groups

```{r, fig.show='hold'}
genes_to_plot <- c("CD19", "CD74","SDC1", "EBF1","PTPRC","CD38","CD24A","CD1D1","CD5","MS4A1","CXCR5","SELL","CD40","CD83","H2-AB1","H2-EB1","CD27","POU2AF1","NT5E","FASL","PDCD1LG2","PRDM1","ITGAM","TGFB1","ZCWPW1")

linesh <- c(1:length(genes_to_plot)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_b[[2]], genes = genes_to_plot, group.by = "group_id", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 45, hjust = 1,vjust = 1), axis.title.x.bottom = element_blank(), axis.ticks = element_line(size = 1), axis.ticks.length = unit(0.15,"cm"))  +scale_size_binned(range = c(0.1,12), n.breaks = 5)  + geom_hline(yintercept = linesh, col = "black", size = 0.9) + scale_color_viridis_c(option = "B", end = 0.85)

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Bc_dottile_group.svg", dpi = 400, width = 5.5, height = 7)
```

## B cells integration with aged CNS dataset

As part of an earlier study, our group examined phenotypes of B cells in the CNS of aged mice. They found increased numbers of lymphocytes and increased clonal expansion, especially of CD8+ B cells in aged vs. young animals. 

(Yermanos, A. et al. Single-cell immune repertoire and transcriptome sequencing reveals that clonally expanded and transcriptionally distinct lymphocytes populate the aged central nervous system in mice. Proc. R. Soc. B Biol. Sci. 288, (2021).)

We wondered how different cells of EAE and aged CNS are transcriptionally. 

Running a new VGM to integrate these datasets. 

```{r, fig.show='hold'}
VDJ.out.directory.list <- list()
VDJ.out.directory.list[[1]] <- c("~/EAE/bcell/S1") #rMOG 
VDJ.out.directory.list[[2]] <- c("~/EAE/bcell/S1") #rMOG
VDJ.out.directory.list[[3]] <- c("~/EAE/bcell/S1") #MOG35-55
VDJ.out.directory.list[[4]] <- c("~/Aged_CNS/vdj/S5") # 3 Months pool
VDJ.out.directory.list[[5]] <- c("~/Aged_CNS/vdj/S6") # 12 months pool
VDJ.out.directory.list[[6]] <- c("~/Aged_CNS/vdj/S7") # 18 months pool
VDJ.out.directory.list[[7]] <- c("~/Aged_CNS/vdj/S8") # 18 months alone
```

```{r, include = F}

VDJ.out.directory.list <- list()
#EAE
VDJ.out.directory.list[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S1") #Protein acute brain + spinal pooled
VDJ.out.directory.list[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S2") #Protein acute brain + spinal pooled
VDJ.out.directory.list[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S3") #Peptide acute brain + spinal
VDJ.out.directory.list[[4]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S5") #3 months old pooled
VDJ.out.directory.list[[5]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S6") #12 months old pooled
VDJ.out.directory.list[[6]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S7") #18 months old pooled
VDJ.out.directory.list[[7]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S8") #18 months old single

GEX.out.directory.list <- list()
#EAE
GEX.out.directory.list[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S1") #Protein acute brain + spinal pooled
GEX.out.directory.list[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S2") #Protein acute brain + spinal pooled
GEX.out.directory.list[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/gex/S3") #Peptide acute brain + spinal
GEX.out.directory.list[[4]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S1") #3 months old pooled
GEX.out.directory.list[[5]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S2") #12 months old pooled
GEX.out.directory.list[[6]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S3") #18 months old pooled
GEX.out.directory.list[[7]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/gex/S4") #18 months old single

vgm_EAEaged <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_EAEaged_Bcs.rds")

```

Run VGM

```{r, fig.show='hold'}

#vgm_EAEaged <- VDJ_GEX_matrix(VDJ.out.directory.list = VDJ.out.directory.list,  
#                              GEX.out.directory.list = GEX.out.directory.list,
#                             exclude.on.cell.state.markers = c("CD3G+", "CD3E+", "CD4+", "CD8A+","CD8B1+","C1QA+","C1QB+","RORA+","NKG7+"), 
#                             group.id = c("EAE", "EAE", "EAE", "aged CNS", "aged CNS", "aged CNS", "aged CNS")) 

#saveRDS(vgm_EAEaged,"EAE_vgm_EAEaged_Bcs.rds")


sample_ids <- c("s1","s2","s3","s4","s5","s6", "s7")
short_replace_ids <- c("rMOG_1", "rMOG_2","MOG35-55", "3m", "12m", "18m", "18m single")

for (i in seq_along(sample_ids)){
  vgm_EAEaged[[2]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_EAEaged[[2]]$sample_id)
}
vgm_EAEaged[[2]]$sample_id <- ordered(as.factor(vgm_EAEaged[[2]]$sample_id), levels = unique(short_replace_ids))

unique(vgm_EAEaged[[2]]$sample_id)

samples_color_aged <- c(samples_color,"#71716F", "#577590", "#B2B7A4", "#D8DAD3")

short_replace_ids <- ordered(as.factor(short_replace_ids), levels = unique(short_replace_ids))
test_col <- data.frame(pal = samples_color_aged, nam = unique(short_replace_ids) , val = runif(7,8,12))

#Color test ggplot
ggplot(test_col, aes(x = nam, y = val, fill= nam)) + geom_bar(stat = "identity") + scale_fill_manual(values = samples_color_aged)

#Nr. cells per sample 
table(vgm_EAEaged[[2]]$sample_id)
```

Next: umap

```{r, fig.show='hold'}
Seurat::DimPlot(vgm_EAEaged[[2]],reduction = "umap", group.by = "sample_id", cols = samples_color_aged, shuffle = T) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold"), legend.text = element_text(size = 15)) + labs(color = "Sample", x = "UMAP1", y = "UMAP2") + guides(colour = guide_legend(override.aes = list(size=5)))

#ggsave(plot = last_plot(), filename = "Bcs_eaeaged_umap_sample.svg", dpi = 700, width = 6.5, height = 5)

Seurat::DimPlot(vgm_EAEaged[[2]],reduction = "umap", group.by = "seurat_clusters", cols = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)], shuffle = T, label =T, label.size = 7 ) + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold"), legend.text = element_text(size = 15)) + labs(color = "Cluster", x = "UMAP1", y = "UMAP2")+ guides(colour = guide_legend(override.aes = list(size=5)))

#ggsave(plot = last_plot(), filename = "Bcs_eaeaged_umap_cluster.svg", dpi = 700, width = 6.5, height = 5)
```

Next we look at cluster membership using a barplot

```{r, fig.show='hold'}
cl_members <- GEX_proportions_barplot(GEX = vgm_EAEaged[[2]],source.group = "sample_id", target.group = "seurat_clusters", stacked.plot = T)+ scale_fill_manual(values = samples_color_aged) + theme(plot.margin = margin(t = 0.7,r = 2.5, unit = "cm"),panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 1), axis.ticks = element_line(size = 1), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "none", plot.title = element_blank(), legend.title = element_text(size = 25, face = "bold"), legend.text = element_text(size = 19)) + labs(title = "", x = "", y = "% of cells of cluster", fill = "Sample") + scale_y_continuous(expand = c(0,0))
cl_members 

#ggsave(last_plot(), filename = "Bcs_eaeaged_Cluster_membership.svg", dpi = 400, width = 7.5, height = 5)
```

Viewing selected genes as a dotmap 

```{r, fig.show='hold'}
genes_to_plot <- c("CD19","CD74","SDC1", "EBF1","PTPRC","CD38","CD24A","CD1D1","CD5","MS4A1","CXCR5","SELL","CD40","CD83","H2-AB1","H2-EB1","CD27","POU2AF1","NT5E","FASL","PDCD1LG2","PRDM1","ITGAM","TGFB1","ZCWPW1")

linesh <- c(1:length(genes_to_plot)) + 0.5
plot_out <- GEX_dottile_plot(GEX = vgm_EAEaged[[2]], genes = genes_to_plot, group.by = "sample_id", threshold.to.plot = 3) + labs(size = "% positive", color = "Expression", x = "Cluster") + theme(plot.title = element_blank(), axis.title.x = element_text(), axis.text.x = element_text(size = 29, angle = 45, hjust = 1,vjust = 1), axis.title.x.bottom = element_blank(), axis.ticks = element_line(size = 1), axis.ticks.length = unit(0.15,"cm"))  +scale_size_binned(range = c(0.1,12), n.breaks = 5)  + geom_hline(yintercept = linesh, col = "black", size = 0.9) + scale_color_viridis_c(option = "B", end = 0.85)

#For vignette
plot_out + theme(legend.position= "none")

#ggsave(plot = plot_out, filename = "Bc_eaeaged_dottile.svg", dpi = 400, width = 8, height = 11)
```

Looking at differential expressed genes by condition. 

```{r, fig.show='hold'}

sample_DEGs <- GEX_DEgenes(vgm_EAEaged[[2]], min.pct = 0.25, grouping.column = "group_id", group1 = "EAE", group2 = "aged CNS", return.plot = T, logFC = F)

#plot volcano
plot_out <- ggplot(sample_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val_adj), col = avg_logFC)) + geom_point(show.legend = F, size = 4, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 1), axis.ticks = element_line(size = 1, color = "black"), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(adj p)",title = "EAE vs. Aged CNS") + geom_text_repel(data = subset(sample_DEGs[[1]], -log10(p_val_adj) > 20 | SYMBOL %in% c("CD19", "EBF1", "CD40", "FASL")), aes(x = avg_logFC, y = -log10(p_val_adj), label = SYMBOL, col = avg_logFC), inherit.aes = F, size = 7, segment.alpha = 1, max.overlaps = 40)+scale_color_gradient2(low = samples_color_aged[4], mid = "grey60", high = samples_color[2], midpoint = 0)

plot_out

#ggsave(plot_out, filename = "Bc_eaeaged_volc.svg", dpi = 400, width = 8, height = 9.2)
```

## B cells VDJ 

After having explored transcriptional phenotypes of B cells we now turn to their TCR repertoires. 
We first look at basic metrics of expansion by plotting clonal donuts 
With VGM counts

```{r, fig.show='hold'}
plots_out <- VDJ_clonal_donut(vgm_b[[1]], counts.to.use = "VGM", expanded.colors = c("#68228B","#BF3EFF","#E066FF","#8968CD"), non.expanded.color = "#000000", label.size = 9, total.label.vjust = 2) 
#here we use counts of clones from the VGM (i.e. the nr. of rows/cells of every clone) This can also be set to "10x" to use counts from the clonotype_frequency column based on original Cellranger output. The two may diverge if a large number of overlapping barcodes were filtered out or if the clonotyping strategy was changed 

for(i in 1:length(plots_out)){ #saving in a loop
  print(plots_out[[i]])
  #ggsave(plots_out[[i]], filename = paste0("Donut_VGMcounts",plots_out[[i]]$labels$title ,".png"), dpi = 400, width = 5, height = 5, bg = "transparent")
}
```

With 10x cellranger counts

```{r, fig.show='hold'}
plots_out <- VDJ_clonal_donut(vgm_b[[1]], counts.to.use = "10x", expanded.colors = c("#68228B","#BF3EFF","#E066FF","#8968CD"), non.expanded.color = "#000000", label.size = 9, total.label.vjust = 2) 

for(i in 1:length(plots_out)){ #saving in a loop
  print(plots_out[[i]])
  #ggsave(plots_out[[i]], filename = paste0("Donut_10xcounts",plots_out[[i]]$labels$title ,".png"), dpi = 400, width = 5, height = 5, bg = "transparent")
}
```

We continue by calculating common diversity metrics
Here we use VDJ and VJ CDR3s seqs as a basis

```{r, fig.show='hold'}
#NO SUBSAMPLING
out_list <- list() #loop to compile results for each metric
for(i in c("richness","bergerparker","ginisimpson","shannonevenness")){
  out <- VDJ_diversity(VDJ = vgm_b[[1]],feature.columns = c("VDJ_cdr3s_aa","VJ_cdr3s_aa"),grouping.column = "sample_id", metric = i, platypus.version = "v3", subsample.to.same.n = F)$data
  out$metric_used <- i
  out_list[[i]] <- out
}
out <- bind_rows(out_list) #generating a common dataframe

#lets plot this
out$index <- ordered(as.factor(out$index), levels = c("Species richness", "Berger-Parker index", "Gini-Simpson index", "Shannon evenness"))
plot_out <- ggplot(out, aes(x = groups, y = metric, fill = groups))  + labs(title = "", x = "", y = "") + theme(panel.background = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") + geom_bar(width = 0.6,show.legend = F, stat = "identity") + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 30),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), strip.background = element_rect(fill = "white", color = "white")) + scale_y_continuous(expand = c(0,0))+ scale_fill_manual(values= samples_color)+ facet_wrap(~index, scales = "free") 

plot_out

#ggsave(plot_out, filename = "Bcs_rep_diversity_nosubsampling.svg", dpi = 600, width = 10, height = 8)

#SUBSAMPLING
out_list <- list()
for(i in c("richness","bergerparker","ginisimpson", "shannonevenness")){
  out <- VDJ_diversity(VDJ = vgm_b[[1]],feature.columns = c("VDJ_cdr3s_aa","VJ_cdr3s_aa"),grouping.column = "sample_id",metric = i, platypus.version = "v3", subsample.to.same.n = T)$data #subsampling down to the smallest repertoire
  out$metric_used <- i
  out_list[[i]] <- out
}
out <- bind_rows(out_list)

#lets plot this too
out$index <- ordered(as.factor(out$index), levels = c("Species richness", "Berger-Parker index", "Gini-Simpson index", "Shannon evenness"))
plot_out <- ggplot(out, aes(x = groups, y = metric, fill = groups))  + labs(title = "", x = "", y = "") + theme(panel.background = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") + geom_bar(width = 0.6,show.legend = F, stat = "identity") + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 30),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), strip.background = element_rect(fill = "white", color = "white")) + scale_y_continuous(expand = c(0,0))+ scale_fill_manual(values= samples_color)+ facet_wrap(~index, ncol = 2, scales = "free") 

plot_out

#ggsave(plot_out, filename = "Bcs_rep_diversity_subsampled.svg", dpi = 600, width = 8, height = 10)
```

V gene usage is an important indicator of repertoire similarity or dissimilarity. 
1. stacked barplot 

```{r, fig.show = 'hold'}
levels(vgm_b[[1]]$sample_id)

EAE_Vgene_usage_barplot_b <- VDJ_Vgene_usage_stacked_barplot(vgm_b[[1]], HC.gene.number = 12, LC.Vgene = F, LC.gene.number = 12, platypus.version = "v3")

#ggplot magic
plot_out <- EAE_Vgene_usage_barplot_b + theme(panel.background = element_blank(),axis.text = element_text(size = 25), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=28),legend.position = "right",axis.text.x = element_text(angle = 30,vjust = 1, hjust=1), plot.title = element_blank(), axis.title.x = element_blank(), plot.margin = margin(t = 0.7, unit = "cm"), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),panel.border = element_blank(), legend.title = element_text(face = "bold"))+ scale_fill_manual(values = pals::stepped2(20))+ scale_x_discrete(breaks=c("1","2","3"),labels=c("rMOG_1", "rMOG_2", "MOG35-55")) + labs(y = "relative usage", fill = "VDJ Vgene")

plot_out

#ggsave(plot = plot_out, filename = paste0("Bcs_VDJ_vgene_by_sample.svg"), dpi = 400, width = 8, height = 7)

#VJ
EAE_Vgene_usage_barplot_b <- VDJ_Vgene_usage_stacked_barplot(vgm_b[[1]], HC.gene.number = 12, LC.Vgene = T, LC.gene.number = 12, platypus.version = "v3")

plot_out <- EAE_Vgene_usage_barplot_b + theme(panel.background = element_blank(),axis.text = element_text(size = 25), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=28),legend.position = "right",axis.text.x = element_text(angle = 30,vjust = 1, hjust=1), plot.title = element_blank(), axis.title.x = element_blank(), plot.margin = margin(t = 0.7, unit = "cm"), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),panel.border = element_blank(), legend.title = element_text(face = "bold"))+ scale_fill_manual(values = pals::stepped2(20))+ scale_x_discrete(breaks=c("1","2", "3"),labels=c("rMOG_1", "rMOG_2", "MOG35-55")) + labs(y = "relative usage", fill = "VJ Vgene") 

plot_out

#ggsave(plot = plot_out, filename = paste0("Bcs_VJ_vgene_by_sample.svg"), dpi = 400, width = 8, height = 7)
```

2. Using Circos plots 
These will generate PDF files, which are not contained within this vignette

```{r, fig.show = 'hold'}
pdf(file = "Bcs_circos_rMOG_1.pdf", onefile = T, width = 6, height = 6)
for_cir <- list(subset(vgm_b[[1]], sample_id == "rMOG_1"))
for_cir[[1]]$sample_id <- as.character(for_cir[[1]]$sample_id)
circos <- VDJ_VJ_usage_circos(VDJ = for_cir[[1]], A.or.B = "both", label.threshold = 5, cell.level = T, clonotype.per.gene.threshold = 10, c.count = T, platypus.version = "v3", filter1H1L = T)
dev.off()

pdf(file = "Bcs_circos_rMOG_2.pdf", onefile = T, width = 6, height = 6)
for_cir <- list(subset(vgm_b[[1]], sample_id == "rMOG_2"))
for_cir[[1]]$sample_id <- as.character(for_cir[[1]]$sample_id)
circos <- VDJ_VJ_usage_circos(VDJ = for_cir[[1]], A.or.B = "both", label.threshold = 5, cell.level = T, clonotype.per.gene.threshold = 10, c.count = T, platypus.version = "v3", filter1H1L = T)
dev.off()

pdf(file = "Bcs_circos_MOG3555.pdf", onefile = T, width = 6, height = 6)
for_cir <- list(subset(vgm_b[[1]], sample_id == "MOG35-55"))
for_cir[[1]]$sample_id <- as.character(for_cir[[1]]$sample_id)
circos <- VDJ_VJ_usage_circos(VDJ = for_cir[[1]], A.or.B = "both", label.threshold = 5, cell.level = T, clonotype.per.gene.threshold = 10, c.count = T, platypus.version = "v3", filter1H1L = T)
dev.off()
```

Public clones are of great interest here and so we assay them together with overlap of other repertoire elements

```{r, fig.show='hold'}

#using the unfiltered (VDJ comb here. All cells, also those with less or more than 1VDJ 1VJ are included)
VDJv_overlap <- VDJ_overlap_heatmap(vgm_b[[1]],feature.columns = c("VDJ_vgene") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3", add.barcode.table = T)
length(unique(vgm_b[[1]]$VDJ_vgene))

VJv_overlap <- VDJ_overlap_heatmap(vgm_b[[1]],feature.columns = c("VJ_vgene") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3", add.barcode.table = T)
length(unique(vgm_b[[1]]$VJ_vgene))

VJ_overlap <- VDJ_overlap_heatmap(vgm_b[[1]],feature.columns = c("VJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3")
length(unique(vgm_b[[1]]$VJ_cdr3s_aa))

VDJ_overlap <- VDJ_overlap_heatmap(vgm_b[[1]],feature.columns = c("VDJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3")
length(unique(vgm_b[[1]]$VDJ_cdr3s_aa))

clone_overlap <- VDJ_overlap_heatmap(vgm_b[[1]],feature.columns = c("VJ_cdr3s_aa","VDJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3")
length(unique(vgm_b[[1]]$pasted))


#make a table out of this for barplots (The numbers are from the length(unique())) prints above) 
to_plot <- data.frame(var = c("VDJ vgene","VJ vgene", "VDJ CDR3", "VJ CDR3", "VJ&VDJ CDR3"), ov = c(29,73,19,45,4), total = c(96,314,1546,1629,2094))
to_plot$fraction <- paste0(round((to_plot$ov / to_plot$total * 100),2),"%")

to_plot$var <- ordered(as.factor(to_plot$var),levels = c("VDJ vgene","VJ vgene", "VDJ CDR3","VJ CDR3", "VJ&VDJ CDR3"))

plot_out <- ggplot(data = to_plot, aes(x = var, y = ov, fill = ov)) + geom_bar(alpha = 1, stat = "identity") + theme(panel.background = element_blank(),axis.text.x = element_text(angle = 30,vjust = 1, hjust=1, size = 30),axis.text.y = element_text(size = 27), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2),  axis.line = element_line(size = 2), axis.ticks = element_line(size = 2, color = "black"), axis.ticks.length = unit(0.25,"cm"), text = element_text(size=28), legend.position = "none", plot.margin = margin(l = 0.5, unit = "cm")) + labs(title = "Repertoire overlap", x = "", y = "shared elements") + geom_text(aes(label = fraction , vjust = -0.5), size = 9)+ scale_y_continuous(expand = c(0,0), limits = c(0,83))  + scale_fill_gradient(low = "grey80", high = stepped2(15)[13], oob = squish_infinite)

plot_out

#ggsave(plot_out, filename = "Bcs_rep_overlap_new.svg", dpi = 400, width = 8, height = 10)

```

To get the table of overlapping clones

```{r, fig.show='hold'}
print(head(clone_overlap[[3]]))
```

Here we also use 3 other datasets that were recently aquired in our group. 
The aged CNS dataset, which has been used extensively above and paired BCR/GEX seq of cells from young and aged murine spleen and bone marrow (BM)
We import, run the VGM and check for overlapping clones

```{r, fig.show='hold'}
VDJ.out.directory.list <- list()
VDJ.out.directory.list[[1]] <- c("~/EAE/bcell/S1") #rMOG 
VDJ.out.directory.list[[2]] <- c("~/EAE/bcell/S1") #rMOG
VDJ.out.directory.list[[3]] <- c("~/EAE/bcell/S1") #MOG35-55
VDJ.out.directory.list[[4]] <- c("~/Aged_organs/vdj/S11") #S11 - old BM vdj
VDJ.out.directory.list[[4]] <- c("~/Aged_organs/vdj/S14") #S14 - old spleen vdj
VDJ.out.directory.list[[6]] <- c("~/Aged_organs/vdj/S12") #S12 - young BM vdj
VDJ.out.directory.list[[7]] <- c("~/Aged_organs/vdj/S13") #S13 - young spleen vdj
VDJ.out.directory.list[[8]] <- c("~/Aged_CNS/vdj/S5") # 3 Months pool
VDJ.out.directory.list[[9]] <- c("~/Aged_CNS/vdj/S6") # 12 months pool
VDJ.out.directory.list[[10]] <- c("~/Aged_CNS/vdj/S7") # 18 months pool
VDJ.out.directory.list[[11]] <- c("~/Aged_CNS/vdj/S8") # 18 months alone
```

```{r, include = F}

vgm_EAEaged <- NULL
VDJv_overlap <- NULL
VJ_overlap <- NULL
VJv_overlap <- NULL
VDJ_overlap <- NULL
clone_overlatp <- NULL
gc()

VDJ.out.directory.list <- list()
#EAE
VDJ.out.directory.list[[1]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S1") #Protein acute brain + spinal pooled
VDJ.out.directory.list[[2]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S2") #Protein acute brain + spinal pooled
VDJ.out.directory.list[[3]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/EAE_n3/bcell/S3") #Peptide acute brain + spinal
#Old organs
VDJ.out.directory.list[[4]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_organs/vdj/S11") #S11 - old BM vdj
VDJ.out.directory.list[[5]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_organs/vdj/S14") #S14 - old spleen vdj
VDJ.out.directory.list[[6]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_organs/vdj/S12") #S12 - young BM vdj
VDJ.out.directory.list[[7]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_organs/vdj/S13") #S13 - young spleen vdj
#Aged CNS
VDJ.out.directory.list[[8]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S5") # 3 Months pool
VDJ.out.directory.list[[9]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S6") # 12 months pool
VDJ.out.directory.list[[10]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S7") # 18 months pool
VDJ.out.directory.list[[11]] <- c("C:/Dokumente usw/Master/Reddy Lab/1_thesis/2_raw data/Aged/Aged_CNS/vdj/S8") # 18 months alone

```

Run VGM with only VDJ loading

```{r, fig.show='hold'}
#vgm_int <- VDJ_GEX_matrix(VDJ.out.directory.list = VDJ.out.directory.list, VDJ.combine = T, filter.overlapping.barcodes.VDJ = T, get.VDJ.stats = T, parallel.processing = "parlapply", trim.and.align = F, group.id = c("EAE","EAE","EAE","Aged organs","Aged organs","Aged organs","Aged organs","Aged CNS","Aged CNS","Aged CNS","Aged CNS"))

#saveRDS(vgm_int, "EAE_vgm_int.rds")
vgm_int <- readRDS("C:/Dokumente usw/Master/Reddy Lab/1_thesis/3_code/4_DB project/EAE website vignette/EAE_vgm_int.rds")

vgm_int[[1]]$sample_id <- paste0(vgm_int[[1]]$sample_id, "s") #for easier Gsubs
#add complete sample names
sample_ids <- paste0("s", c(1:11), "s") 
short_replace_ids <- c("rMOG_1", "rMOG_2" , "MOG35-55", "Aged BM", "Aged spleen", "Young BM", "Young spleen", "3m CNS", "12m CNS", "18m CNS", "18m single CNS")

for (i in 1:length(sample_ids)){
  vgm_int[[1]]$sample_id <- gsub(sample_ids[i], short_replace_ids[i], vgm_int[[1]]$sample_id)}
vgm_int[[1]]$sample_id <- ordered(as.factor(vgm_int[[1]]$sample_id), levels = short_replace_ids)

table(vgm_int[[1]]$sample_id)
```

Find overlapping clones

```{r, fig.show='hold'}
clone_overlap <- VDJ_overlap_heatmap(vgm_int[[1]],feature.columns = c("VJ_cdr3s_aa","VDJ_cdr3s_aa") ,grouping.column = "sample_id", axis.label.size = 20, pvalues.label.size = 12, platypus.version = "v3", plot.type = "pheatmap")
length(unique(vgm_int[[1]]$pasted))

clone_overlap[[1]] 

print(head(clone_overlap[[2]][[1]])) #table of overlapping elements
```

To visualize clone similarity across repertoires we display a similarity graph depicting clonal relationships based on the edit distance of CDR3s.

```{r, fig.show = 'hold'}
#get unique clones to avoid cluttering 
vgm_b[[1]]$sample_clone <- paste0(vgm_b[[1]]$sample_id, vgm_b[[1]]$clonotype_id_10x)
vgm_b_unique_clones <- subset(vgm_b[[1]][which(!duplicated(vgm_b[[1]]$sample_clone)),], clonotype_frequency > 2) #filtering: only one cell of each clone with a freq > 10
vgm_b_unique_clones$group_id <- as.character(vgm_b_unique_clones$group_id)
vgm_b_unique_clones$group_id[vgm_b_unique_clones$sample_id == "rMOG_1"] <- samples_color[1] #renewing the group id with the respective sample colors
vgm_b_unique_clones$group_id[vgm_b_unique_clones$sample_id == "rMOG_2"] <- samples_color[2]
vgm_b_unique_clones$group_id[vgm_b_unique_clones$sample_id == "MOG35-55"] <- samples_color[3]

#make a loop to plot for each sample a range of distance cutoff values
pdf(file = "Bcs_VDJ_network_filtered.pdf")
for(j in 3:12){ #generates a plot with different edit distance cutoffs to explore how the network behaves 
  netwo <- VDJ_network(VDJ = vgm_b_unique_clones,distance.cutoff = j,per.sample = F,platypus.version = "v3")
  igraph::plot.igraph(netwo[[4]],vertex.label=NA,vertex.size=5+(.03*as.numeric(netwo[[2]]$clonotype_frequency)),vertex.color=netwo[[2]]$group_id)
}
dev.off()

#for the purpose of this document, we view the plot at j = 10
netwo <- VDJ_network(VDJ = vgm_b_unique_clones,distance.cutoff = 10,per.sample = F,platypus.version = "v3")
igraph::plot.igraph(netwo[[4]],vertex.label=NA,vertex.size=5+(.03*as.numeric(netwo[[2]]$clonotype_frequency)),vertex.color=netwo[[2]]$group_id)
```

## B cell Isotypes

Isotypes are an important metric of any B cell repertoire 
First we examine clonal expansion

```{r, fig.show = 'hold'}
clonal_out <- VDJ_clonal_expansion(VDJ = vgm_b[[1]],celltype = "Bcells",clones = 30, platypus.version = "v3", group.by = "sample_id", color.by = "isotype", subtypes = F, treat.incomplete.clones = "exclude", treat.incomplete.cells = "proportional")

isotypes_color <- c(pals::kelly(n = 14)[3],"darkorchid3" ,pals::kelly(n = 14)[5],"chartreuse3","black","grey60")

#rMOG_1
clonal_out[[1]][[1]] <- clonal_out[[1]][[1]]+ theme(panel.background = element_blank(),axis.text.x = element_text(size = 25),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = c(0.4, 0.9),legend.text = element_text(size = 24), plot.subtitle = element_blank(), plot.title = element_blank(), legend.title = element_text(size = 26, face = "bold")) + scale_fill_manual(values = c(isotypes_color, "grey60")) + labs(fill = "Isotype", y = "n cells")+ ggplot2::scale_y_continuous(expand = c(0,0), breaks = c(2,10,20,30,40,50))+guides(fill=guide_legend(nrow=2,byrow=TRUE)) + geom_hline(yintercept = 2, size = 1, color = "grey60") #adding a horizonal line at freq = 2
clonal_out[[1]][[1]]

#ggsave(plot = clonal_out[[1]][[1]], filename = paste0("Bcs1_expansion_by_isotype_rMOG_1.svg"), dpi = 400, width = 10, height = 8)

#rMOG_2
clonal_out[[1]][[2]] <- clonal_out[[1]][[2]]+ theme(panel.background = element_blank(),axis.text.x = element_text(size = 25),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = c(0.4, 0.9),legend.text = element_text(size = 24), plot.subtitle = element_blank(), plot.title = element_blank(), legend.title = element_text(size = 26, face = "bold")) + scale_fill_manual(values = c(isotypes_color, "grey60")) + labs(fill = "Isotype", y = "n cells")+ ggplot2::scale_y_continuous(expand = c(0,0), breaks = c(2,10,20,30,40,50))+guides(fill=guide_legend(nrow=2,byrow=TRUE)) + geom_hline(yintercept = 2, size = 1, color = "grey60") #adding a horizonal line at freq = 2
clonal_out[[1]][[2]]

#ggsave(plot = clonal_out[[1]][[2]], filename = paste0("Bcs1_expansion_by_isotype_rMOG_1.svg"), dpi = 400, width = 10, height = 8)

#MOG35-55
clonal_out[[1]][[3]] <- clonal_out[[1]][[3]]+ theme(panel.background = element_blank(),axis.text.x = element_text(size = 25),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none",legend.text = element_text(size = 24), plot.subtitle = element_blank(), plot.title = element_blank(), legend.title = element_text(size = 26, face = "bold")) + scale_fill_manual(values = c(isotypes_color, "grey60")) + labs(fill = "Isotype", y = "n cells")+ ggplot2::scale_y_continuous(expand = c(0,0), breaks = c(2, 5,10,15,20))+guides(fill=guide_legend(nrow=3,byrow=TRUE))+ geom_hline(yintercept = 2, size = 1, color = "grey60")
clonal_out[[1]][[3]]

#ggsave(plot = clonal_out[[1]][[3]], filename = paste0("Bcs1_expansion_by_isotype_MOG35-55.svg"), dpi = 400, width = 10, height = 8)
```

Comparing isotype usage between samples

```{r, fig.show = 'hold'}
#we format a column so to make the annotation more consistent
vgm_b[[2]]$isotype <- substr(vgm_b[[2]]$VDJ_cgene, 1,4) #First 4 letters of the c gene
vgm_b[[2]]$isotype[str_detect(vgm_b[[2]]$VDJ_cgene, ";")] <- "Multiple VDJ chains"
vgm_b[[2]]$isotype[vgm_b[[2]]$VDJ_cgene == ""] <- "Unknown"
vgm_b[[2]]$isotype[is.na(vgm_b[[2]]$VDJ_cgene)] <- "Unknown"

vgm_b[[2]]$isotype <- ordered(as.factor(vgm_b[[2]]$isotype), levels = c("IGHA", "IGHG", "IGHD", "IGHM", "IGHE", "Multiple VDJ chains", "Unknown"))
unique(vgm_b[[2]]$isotype)     

plot_out <- GEX_proportions_barplot(GEX = vgm_b[[2]], source.group = "isotype", target.group = "seurat_clusters", stacked.plot = T)

plot_out <-  plot_out + scale_fill_manual(values = c(pals::kelly(n = 14)[3],"darkorchid3" ,pals::kelly(n = 14)[5],"chartreuse3","black","grey60","grey80")) + theme(panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "right", plot.title = element_blank(), legend.title = element_text(size = 30, face = "bold")) + labs(y = "% of clonal level", x = "Cluster", fill = "Clonal level")

plot_out
  
#ggsave(plot = plot_out, filename = paste0("Bcs_prop_isotype.svg"), dpi = 700, width = 8, height = 6) 

plot_out <- plot_out + theme(legend.position = "none")
#ggsave(plot = plot_out, filename = paste0("Bcs_prop_isotype_noleg.svg"), dpi = 700, width = 8, height = 6) 
```

we will return to isotypes upon VDJ GEX integration

## B cells SHM

To quantify somatic hypermutations we use MIXCR and align BCR sequences to then recover the count of inserts, deletions and mismatches. Using MIXCR requires that the user has downloaded the software and can locate the correct directory.

```{r, fig.show='hold', eval = TRUE, include=FALSE}
vgm_b_bk <- vgm_b
vgm_b <- vgm_b_bk
```

```{r, fig.show='hold', message=FALSE, echo=FALSE, prompt=FALSE, results='hide'}
#add columns directly to the VGM
vgm_b[[1]] <- VDJ_call_MIXCR(VDJ = vgm_b[[1]], operating.system = "Windows", mixcr.directory = "", species = "mmu", simplify = T, platypus.version = "v3")
#Mixcr directory left blank as it is anyways set to the current working directory (getwd()) on a windows machine. 
#Simplify = T to reduce the number of added columns
```

A column containing SHM information for VDJ and VJ chain has now been added, which can be used directly for plotting 

```{r, fig.show='hold'}
names(vgm_b[[1]])
```

Run the function, which returns multiple plots, then take view the by sample boxplots
This will also return the results of signifcance analysis by anova and post hoc tukey HSD tests. 

```{r, fig.show='hold'}
SHM_plots <- VDJ_plot_SHM(VDJ = vgm_b[[1]], 
                     group.by = "sample_id", 
                     quantile.label = 0.995,
                     point.size = 2.7,
                     platypus.version = "v3")

#Boxplot
plot_out <- SHM_plots[[1]] + theme(plot.margin = margin(0, 0, 0, 2, "cm"), panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"), axis.text.x = element_text(angle = 50,vjust = 1, hjust=1), legend.position = "none") + scale_color_manual(values = samples_color)
  
plot_out

#ggsave(plot_out,filename = "Bcs_SHM_box_sample.svg",dpi = 600, width = 8.5, height = 11) 
```

Of course it is also interesting to view VGM by isotype

```{r, fig.show='hold'}

#adding formatted isotype column to VDJ
vgm_b[[1]]$isotype <- substr(vgm_b[[1]]$VDJ_cgene, 1,4) #First 4 letters of the c gene
vgm_b[[1]]$isotype[str_detect(vgm_b[[1]]$VDJ_cgene, ";")] <- "Multiple VDJ chains"
vgm_b[[1]]$isotype[vgm_b[[1]]$VDJ_cgene == ""] <- "Unknown"
vgm_b[[1]]$isotype[is.na(vgm_b[[1]]$VDJ_cgene)] <- "Unknown"

SHM_plots <- VDJ_plot_SHM(VDJ = vgm_b[[1]], 
                     group.by = "isotype", 
                     quantile.label = 0.995,
                     point.size = 2.7,
                     platypus.version = "v3")

#Boxplot
plot_out <- SHM_plots[[1]] + theme(plot.margin = margin(0, 0, 0, 2, "cm"), panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"), axis.text.x = element_text(angle = 50,vjust = 1, hjust=1), legend.position = "none") + scale_color_manual(values = c(pals::kelly(n = 14)[3],"darkorchid3" ,pals::kelly(n = 14)[5],"chartreuse3","black","grey60","grey80"))
  
plot_out

#ggsave(plot_out,filename = "Bcs_SHM_box_isotype.svg", dpi = 600, width = 8.5, height = 11) 
```

Finally we also visualize SHM at the clonal level

```{r, fig.show='hold'}
#Adding a dedicated column of sample + clonal level
vgm_b[[1]]$expanded <- "1 cell"
vgm_b[[1]]$expanded[vgm_b[[1]]$clonotype_frequency > 1] <- "Expanded"

vgm_b[[1]]$sample_expanded <- paste0(vgm_b[[1]]$sample_id, " ", vgm_b[[1]]$expanded)

unique(vgm_b[[1]]$sample_expanded)
vgm_b[[1]]$sample_expanded <- ordered(as.factor(vgm_b[[1]]$sample_expanded), levels = c("rMOG_1 Expanded","rMOG_2 Expanded", "MOG35-55 Expanded", "rMOG_1 1 cell", "rMOG_2 1 cell", "MOG35-55 1 cell" ))


SHM_plots <- VDJ_plot_SHM(VDJ = vgm_b[[1]], 
                     group.by = "sample_expanded", 
                     quantile.label = 0.995,
                     point.size = 2.7,
                     platypus.version = "v3")

#Boxplot
plot_out <- SHM_plots[[1]] + theme(plot.margin = margin(0, 0, 0, 2, "cm"), panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"), axis.text.x = element_text(angle = 50,vjust = 1, hjust=1), legend.position = "none") + scale_color_manual(values = c(samples_color,samples_color))
  
plot_out

#ggsave(plot_out,filename = "Bcs_SHM_box_isotype.svg", dpi = 600, width = 8.5, height = 11) 
```

## B cells VDJ GEX integration 

The first step is to check how well these two datasets overlap. 

Nr. of cells in VDJ for which GEX info is available

```{r, fig.show='hold'}
GEX_av <- vgm_b[[1]] %>% group_by(sample_id) %>% summarize(av = sum(GEX_available), not_av = n()-sum(GEX_available)) #Using the sum of this column as in R, TRUE is counted as numeric 1 and FALSE as numeric 0
GEX_av <- pivot_longer(GEX_av, cols = c(2,3))

plot_out <- list()
for(i in 1:3){

GEX_av_plot <- subset(GEX_av, sample_id == unique(GEX_av$sample_id)[i])
  
plot_out[[i]] <- ggplot(GEX_av_plot, aes(x = "", y=value, fill= name))+ geom_bar(stat = "identity", width = 1, show.legend = F)+ scale_fill_manual(values = c("forestgreen","grey60"))+ coord_polar("y", direction = -1)+ theme(panel.background = element_blank(),axis.text.x = element_blank(),axis.text.y = element_blank(), axis.line = element_blank(), axis.ticks = element_blank(), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right",legend.direction = "vertical", plot.title = element_text(size = 20,hjust = 0.5)) + labs(x = "", y = "", title = paste0(unique(GEX_av$sample_id)[i])) + geom_text(x = 1, hjust = 1.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "not_av"], color = "white", fontface = "bold", size = 8) + geom_text(x = 1, hjust = -0.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "av"], color = "white", fontface = "bold", size = 8) + geom_point(inherit.aes = F, aes(x = 0, y = 1), color = "white", size = 25) 

ggsave(plot_out[[i]], filename = paste0("GEX_available_pie_",plot_out[[i]]$labels$title ,".png"), dpi = 600, width = 5, height = 5)
}

print("Purple = Cells in VDJ for which GEX information is available")
print(plot_out)
print(GEX_av)
```

Nr. of cells in GEX for which VDJ info is available

```{r, fig.show='hold'}
VDJ_av <- vgm_b[[2]]@meta.data %>% group_by(sample_id) %>% summarize(av = sum(VDJ_available), not_av = n()-sum(VDJ_available)) 
VDJ_av <- pivot_longer(VDJ_av, cols = c(2,3))

plot_out <- list()
for(i in 1:3){

VDJ_av_plot <- subset(VDJ_av, sample_id == unique(VDJ_av$sample_id)[i])
  
plot_out[[i]] <- ggplot(VDJ_av_plot, aes(x = "", y=value, fill= name))+ geom_bar(stat = "identity", width = 1, show.legend = F)+ scale_fill_manual(values = c("darkorchid3","grey60"))+ coord_polar("y", direction = -1)+ theme(panel.background = element_blank(),axis.text.x = element_blank(),axis.text.y = element_blank(), axis.line = element_blank(), axis.ticks = element_blank(), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right",legend.direction = "vertical", plot.title = element_text(size = 20,hjust = 0.5)) + labs(x = "", y = "", title = paste0(unique(GEX_av$sample_id)[i])) + geom_text(x = 1, hjust = 1.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "not_av"], color = "white", fontface = "bold", size = 8) + geom_text(x = 1, hjust = -0.5, vjust = 0.8, y = 1, label = GEX_av_plot$value[GEX_av_plot$name == "av"], color = "white", fontface = "bold", size = 8) + geom_point(inherit.aes = F, aes(x = 0, y = 1), color = "white", size = 25) 

#ggsave(plot_out[[i]], filename = paste0("VDJ_available_pie_",plot_out[[i]]$labels$title ,".png"), dpi = 600, width = 5, height = 5)

}

print("Purple = Cells in GEX for which VDJ information is available")
print(plot_out)
print(VDJ_av)
```

As a second step we now overlay VDJ information onto the GEX umap

```{r, fig.show='hold'}
#Add a discrete scale for expansion (To plot a continuos scale, use Seurat::FeaturePlot(..., features = "clonotype_frequency"))
vgm_b[[2]]$expanded <- vgm_b[[2]]$clonotype_frequency
vgm_b[[2]]$expanded[vgm_b[[2]]$clonotype_frequency > 1] <- "Expanded"
vgm_b[[2]]$expanded[vgm_b[[2]]$clonotype_frequency == 1] <- "1 cell"
vgm_b[[2]]$expanded[vgm_b[[2]]$VDJ_available == F] <- "VDJ not available"

plot_out <- Seurat::DimPlot(vgm_b[[2]],reduction = "umap", group.by = "expanded", cols = c("grey30","darkorchid3","grey80"), shuffle = T)+ theme(axis.text.y = element_blank(),axis.text.x = element_blank(),axis.line=element_blank(),axis.ticks = element_blank(), legend.position = "right") 
plot_out <- plot_out + theme( axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Clonal level", x = "UMAP1", y = "UMAP2") +guides(color=guide_legend(nrow=3,byrow=TRUE, override.aes = list(size=4)))

plot_out 

#ggsave(plot = plot_out , filename = "Bcs_GEXVDJ_expansion_discrete.svg", dpi = 700, width = 6.8, height = 5) 
```

Proportion barplots again help us visualize differential distribution among clusters

```{r, fig.show='hold'}
#adding a combination column
vgm_b[[2]]$sample_expanded <- paste0(vgm_b[[2]]$expanded, " ", vgm_b[[2]]$sample_id)
vgm_b[[2]]$sample_expanded[str_detect(vgm_b[[2]]$sample_expanded, "VDJ not available")] <- "VDJ not available" #replacing sample specific VDJ not available with a general one (e.g. rMOG_1 VDJ not available -> VDJ not available)
vgm_b[[2]]$sample_expanded <- ordered(as.factor(vgm_b[[2]]$sample_expanded), levels = c("Expanded rMOG_1", "1 cell rMOG_1","Expanded rMOG_2", "1 cell rMOG_2", "Expanded MOG35-55", "1 cell MOG35-55", "VDJ not available"))

plot_out <- GEX_proportions_barplot(GEX = vgm_b[[2]], source.group = "seurat_clusters", target.group = "sample_expanded", stacked.plot = T)

plot_out <-  plot_out + scale_fill_manual(values = pals::kelly(n = 17)[c(2,3,17,7:13,6,16)])  + theme(panel.background = element_blank(),axis.text = element_text(size = 18), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=26), legend.key = element_rect(colour = "white"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), plot.margin = margin(l = 1, t = 1, unit = "cm"), legend.title = element_text(size = 18, face = "bold")) + labs(y = "% of cells of B cell group", x = "", fill = "Cluster")
  
plot_out

#ggsave(plot = plot_out, filename = paste0("Bcs_prop_sampleexpanded.svg"), dpi = 700, width = 10, height = 9) 
```

We are now ready to examine individual clones and their transcriptional phenotypes
The VDJ_clonal_expansion function gives great flexibility.

1. Expansion barplots by cluster (plotting by group ID given the large overlap between rMOG samples)

```{r, fig.show='hold'}
clonal_out <- VDJ_clonal_expansion(VDJ = vgm_b[[1]],celltype = "Bcells",clones = "50", platypus.version = "v3", group.by = "group_id", color.by = "seurat_clusters")

for(i in 1:length(clonal_out[[1]])){ #as this function will return a nested list of source dataframes and plots we loop over it to refine the given plots
  
clonal_out[[1]][[i]] <- clonal_out[[1]][[i]] + theme(panel.background = element_blank(),axis.text.x = element_text(size = 25),axis.text.y = element_text(size = 30), axis.line.x = element_line(size = 2),axis.line.y = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=30), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "bottom", plot.subtitle = element_text(size = 15), legend.title = element_text(size = 26, face = "bold")) + labs(title = unique(vgm_b[[1]]$group_id)[i]) + scale_fill_manual(values = c(pals::kelly(n = 17)[c(2,3,17,7,8)], "grey60","grey40")) + labs(fill = "Cluster", y = "n cells")+ ggplot2::scale_y_continuous(expand = c(0,0))

#ggsave(plot = clonal_out[[1]][[i]], filename = paste0("Bcs_cluster_per_clone_s_", unique(vgm_b[[1]]$group_id)[i],".svg"), dpi = 400, width = 11, height = 7)
}

clonal_out[[1]]
```

Fitting with this we can return to isotypes and view those projected onto the UMAP

```{r, fig.show='hold'}
plot_out <- Seurat::DimPlot(vgm_b[[2]],reduction = "umap", group.by = "isotype", cols = c(pals::kelly(n = 14)[3],"darkorchid3" ,pals::kelly(n = 14)[5],"chartreuse3","black","grey60","grey80"), shuffle = T)+ theme(axis.text.y = element_blank(),axis.text.x = element_blank(),axis.line=element_blank(),axis.ticks = element_blank(), legend.position = "right") 

plot_out <- plot_out + theme(axis.ticks = element_blank(), legend.position = "right",plot.title =element_blank(),axis.line=element_line(size = 1),axis.text = element_blank(), legend.title = element_text(size = 14, face = "bold")) + labs(color = "Isotpye", x = "UMAP1", y = "UMAP2") +guides(color=guide_legend(nrow=8,byrow=TRUE, override.aes = list(size=4)))

plot_out

#ggsave(plot = plot_out , filename = "Bcs_GEXVDJ_umap_isotype.svg", dpi = 700, width = 6.8, height = 5) 
```

Checking the same isotype distribution with a proportions stacked barplot

```{r, fig.show='hold'}
plot_out <- GEX_proportions_barplot(GEX = vgm_b[[2]], source.group = "isotype", target.group = "seurat_clusters", stacked.plot = T)

plot_out <-  plot_out + scale_fill_manual(values = c(pals::kelly(n = 14)[3],"darkorchid3" ,pals::kelly(n = 14)[5],"chartreuse3","black","grey60","grey80")) + theme(panel.background = element_blank(),axis.text = element_text(size = 33), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.25, "cm"), text = element_text(size=33), legend.key = element_rect(colour = "grey"), legend.position = "right", plot.title = element_blank(), legend.title = element_text(size = 30, face = "bold")) + labs(y = "% of isotype", x = "Cluster", fill = "Isotype")

plot_out
  
#ggsave(plot = plot_out, filename = paste0("Bcs2_prop_isotype.svg"), dpi = 700, width = 10, height = 6) 
```

Proceeing to DEG analysis by expansion 

```{r, fig.show='hold'}
exp_DEGs <- GEX_DEgenes(vgm_b[[2]], grouping.column = "expanded", group1 = "Expanded", group2 = "1 cell", min.pct = 0.25, logFC = F)

#Again remove this particular gene that already showed an abnormally high fold change in rMOG vs. MOG35-55
exp_DEGs[[1]] <- exp_DEGs[[1]][rownames(exp_DEGs[[1]]) != "GM30211",]

#given the low statistical power here, we plot non-adjusted p values
plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val), col = avg_logFC)) + geom_point(show.legend = F, size = 3, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.15, "cm"), text = element_text(size=26), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(p)",title = "", subtitle = "Expanded vs. 1 cell") + geom_text_repel(data = subset(exp_DEGs[[1]], p_val < 0.0005 | (avg_logFC > 0 & p_val < 0.05)), aes(x = avg_logFC, y = -log10(p_val), label = SYMBOL), inherit.aes = F, size = 5, segment.alpha = 1, max.overlaps = 50) + scale_colour_gradient2(low = "grey30", mid = "grey60", high = "darkorchid3", oob = squish_infinite)

plot_out

#ggsave(plot_out, filename = paste0("Bcs1_DEGexpanded_pval.svg"), dpi = 400, width = 6.7, height = 6)
```

And to elaborate on the comparison between immunization conditions 

```{r, fig.show='hold'}
#Add a group + expansion column
vgm_b[[2]]$group_expanded <- paste0(vgm_b[[2]]$group_id, " ", vgm_b[[2]]$expanded)
unique(vgm_b[[2]]$group_expanded)

#Expanded vs. Expanded from each group
exp_DEGs <- GEX_DEgenes(vgm_b[[2]], grouping.column = "group_expanded", min.pct = 0.25, group1 = "rMOG Expanded", group2 = "MOG35-55 Expanded", return.plot = "volcano", logFC = F)

exp_DEGs[[1]] <- exp_DEGs[[1]][rownames(exp_DEGs[[1]]) != "GM30211",]

#plot volcano
plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val), col = avg_logFC)) + geom_point(show.legend = F, size = 3, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=26), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(p)",title = "Expanded rMOG vs. MOG35-55", subtitle = "") + geom_text_repel(data = subset(exp_DEGs[[1]], -log10(p_val) > 2), aes(x = avg_logFC, y = -log10(p_val), label = SYMBOL), inherit.aes = F, size = 5, segment.alpha = 1, max.overlaps = 50) + scale_colour_gradient2(low = samples_color[2], mid = "grey60", high = samples_color[1], oob = squish_infinite)

plot_out

#ggsave(plot_out, filename = paste0("Bcs1_DEG_expanded rMOG vs MOG3555.svg"), dpi = 400, width = 6, height = 6)

#now for non expanded cells
exp_DEGs <- GEX_DEgenes(vgm_b[[2]], grouping.column = "group_expanded", min.pct = 0.25, group1 = "rMOG 1 cell", group2 = "MOG35-55 1 cell", return.plot = "volcano", logFC = F)
exp_DEGs[[1]] <- exp_DEGs[[1]][rownames(exp_DEGs[[1]]) != "GM30211",]

#plot volcano
plot_out <- ggplot(exp_DEGs[[1]], aes(x = avg_logFC, y = -log10(p_val), col = avg_logFC)) + geom_point(show.legend = F, size = 3, alpha = 0.6)  + theme(panel.background = element_blank(),axis.text = element_text(size = 30), axis.line = element_line(size = 2), axis.ticks = element_line(size = 2), axis.ticks.length = unit(0.3, "cm"), text = element_text(size=26), legend.key = element_rect(colour = "grey"),legend.key.height = NULL,legend.key.width = NULL, legend.position = "none") + labs(x = "log2(avg FC)", y = "-log10(p)",title = "1 cell rMOG vs. MOG35-55", subtitle = "") + geom_text_repel(data = subset(exp_DEGs[[1]], -log10(p_val) > 5), aes(x = avg_logFC, y = -log10(p_val), label = SYMBOL), inherit.aes = F, size = 4, segment.alpha = 1, max.overlaps = 50) + scale_colour_gradient2(low = samples_color[2], mid = "grey60", high = samples_color[1], oob = squish_infinite)

plot_out

#ggsave(plot_out, filename = paste0("Bcs1_DEG_1cell rMOG vs MOG3555.png"), dpi = 400, width = 6, height = 6)
```

As part of the investigation into BCR specificities within EAE, we selected 20 BCR sequences to express in the available PnP hybridoma system. (Pogson, M., Parola, C., Kelton, W. J., Heuberger, P. & Reddy, S. T. Immunogenomic engineering of a plug-and-(dis)play hybridoma platform. Nat. Commun. 7, 12535 (2016)).
As part of this, aligned BCR sequences need to be prepared for DNA synthesis. 
For this the function VDJ_assemble_for_PnP helps

```{r, fig.show='hold', eval = FALSE}
VGM_with_PnP_seq <- VDJ_assemble_for_PnP(VDJ.mixcr.matrix = vgm_b[[1]], id.column = "barcode",species = "human", manual_IgKC = "none", manual_2A = "none", manual_VDJLeader = "none", write.to.disk = T, filename = "RawforPnP_EAE.csv")
#We use human constant regions so to allow for staining of tissues via microscopy.

print(VGM_with_PnP_seq[1,])
```

```{r, fig.show='hold'}
utils::sessionInfo()
```

